\section{Introduction}
\label{sec:intro}

Kolmogorov's axioms are usually read as fixing the arithmetic of
chance: non-negativity, normalization, additivity.  But they fix
something else first, so quietly that it is easy to miss: a
\emph{logic}.  The events form a Boolean $\sigma$-algebra, closed
under complement and satisfying excluded middle, and the arithmetic
is built on top of that logical substrate.  The complement rule
$P(\neg A) = 1 - P(A)$ is where the two layers touch.  It is an
arithmetical equation about a logical operation.

There is no intrinsic reason why probability should be compatible
with exactly one logic.  Different situations are described by
different logics.  Intuitionistic logic describes computation,
verification, and finite-precision observation, where a proposition
may be confirmable but not refutable.  Linear logic describes
resources, where using an assumption consumes
it~\cite{girard1987}.  Quantum logic describes measurement, where
propositions fail to distribute.  If probability is, as the
Cox--Jaynes tradition has it, \emph{extended
logic}~\cite{cox1946,jaynes2003}, the calculus of degrees of
certainty over a space of propositions, then the logic of those
propositions is a \emph{parameter} of the theory, and Kolmogorov's
calculus is the value of that parameter at Boolean logic.

This is not a speculative proposal, because physics has already
forced the move once.  Quantum probability lives on orthomodular
lattices of projections rather than Boolean algebras, and Gleason's
theorem~\cite{gleason1957,birkhoff-vonneumann1936} plays exactly the
role there that this paper's representation theorems play here:
identify the states compatible with the event logic.  Nobody
concludes that quantum probability is ``not really probability.''
The conclusion is that the Boolean substrate was a modeling choice.
Our question is what happens at a different, computationally
motivated value of the same parameter:

\begin{quote}
  What does probability become when the logic of events is
  \emph{intuitionistic}, and can the resulting theory be developed,
  end to end, under a proof assistant?
\end{quote}

Intuitionistic logic is the natural first non-Boolean instance for a
verification audience.  Its propositions are the \emph{verifiable}
ones: semidecidable properties of programs, finite-precision
observations of a real quantity, opens of a topological space.  Its
algebraic semantics is mature order theory, well supported by
\textsf{mathlib}~\cite{mathlib2020}: frames, that is, complete
Heyting algebras, the lattices of open sets.  And nothing is lost.
Boolean algebras are exactly the frames where excluded middle holds,
so the classical theory embeds as a special case rather than being
replaced.

\paragraph{What changes, in one paragraph.}
A \emph{valuation} on a frame $\Omega$ is a monotone, normalized,
modular map $v : \Omega \to \ENNReal$
(Section~\ref{sec:valuation}).  Modularity is inclusion--exclusion,
$v(a) + v(b) = v(a \sqcup b) + v(a \sqcap b)$, and no axiom mentions
complement.  In a Heyting algebra the pseudo-complement satisfies
$a \sqcap \neg a = \bot$ but not, in general,
$a \sqcup \neg a = \top$.  The complement rule therefore degrades
to a derived \emph{inequality} $v(a) + v(\neg a) \le 1$, with a gap
\[
  \slack(a) \;=\; 1 - v(a) - v(\neg a)
\]
measuring exactly the valuation of the region where excluded middle
fails for $a$.  In the concrete model where $\Omega$ is the opens of
a space carrying a probability measure $\mu$, the slack of an open
$U$ is precisely $\mu(\partial U)$, the measure of its topological
boundary (Section~\ref{sec:bridges}): a structural quantity, not
anyone's ignorance.  The theory that results is recognizably the
\emph{shape} of Dempster--Shafer belief
functions~\cite{dempster1967,shafer1976}, but derived from a
logical substrate rather than postulated, with the
belief/plausibility gap given a non-epistemic reading.

\paragraph{What changes operationally.}
The statistical \emph{operations} are logic-uniform.  The
conditional is still $v(a \sqcap b)/v(b)$, and the product rule and
Bayes symmetry hold verbatim.  What varies is the set of
\emph{identities} those operations satisfy, and with it the licensed
computations.  The classical case split over $\{A, \neg A\}$, and
with it the usual denominator of Bayes' theorem
$P(B) = P(B \mid A)\,P(A) + P(B \mid \neg A)\,P(\neg A)$, degrades to
a lower bound whose deficit is a conditional slack.  Marginalization
is exact only over \emph{genuine} partitions
(Section~\ref{sec:valuation}).  The complement trick
$P(A) = 1 - P(\neg A)$ degrades to the interval
$[\,v(A),\, 1 - v(A^{\mathsf{c}})\,]$ of width $\slack(A)$, so point
statistics becomes interval statistics exactly on the events the
logic leaves undecided, and collapses back to points on complemented
ones.  Updating itself forks.  Dempster's rule and Bayesian
conditioning, classically indistinguishable, give different
posteriors from the same evidence unless its slack vanishes
(Theorem~\ref{thm:cond-hinge}).  That is a modeling choice classical
statistics never faces, because excluded middle collapses the fork.

\paragraph{\ldots and when it does not.}
On finite frames every valuation \emph{is} a classical probability
mass function on points (Theorem~\ref{thm:finite-rep}).  The logic
changes not the underlying chance but which queries have point
values: one computes $P(\Box A)$, the probability of the verifiable
part.  In the measure model slack is boundary mass, so a nonatomic
distribution queried on intervals behaves entirely classically.  The
differences bite precisely at events with charged boundaries and at
semidecidable hypotheses.  The logic sits upstream of statistics the
way a type system sits upstream of programs: the same syntax of
operations, a different set of valid transformations, and some
derivations that typecheck classically (the case split, the
complement trick, ``the'' posterior) either fail or split in two.

\paragraph{Why a proof assistant.}
This subject sits at a triple point of order theory, analysis, and
logic, where informal arguments have a documented history of going
wrong.  Cox's original theorem required repair after Halpern's
counterexample~\cite{cox1946,halpern1999,vanhorn2003}, and the
constructive setting multiplies the opportunities for smuggled
classicality.  Mechanization was not a formality here.  It did real
work.  The checker refuted our first statement of the central
Cox-style theorem, which was not merely unprovable but
\emph{unsatisfiable} (Section~\ref{sec:cox}), and the repair
identified precisely which axiom of the classical development is
irreducible: the sum rule must be posited, because a disjunction's
value is provably not a function of the values of its disjuncts
(Theorem~\ref{thm:no-functional}).  The checker likewise refuted the
folklore guess that the ``classical fragment'' of the theory is the
$\neg\neg$-stable propositions (it is the complemented ones,
Section~\ref{sec:valuation}), and it forced non-vacuity guards:
explicit models witnessing that every axiom set in the paper is
inhabited by genuinely non-classical instances, including one where
collapse to classical logic is impossible \emph{on pain of solving
the halting problem} (Section~\ref{sec:bridges}).

\paragraph{Contributions.}
All results are formalized in Lean~4 over \textsf{mathlib}.  The
development is about 3{,}950 lines across fourteen modules,
\texttt{sorry}-free, with every landmark theorem depending only on
\texttt{propext}, \texttt{Classical.choice}, and \texttt{Quot.sound}
(Section~\ref{sec:engineering}).  To our knowledge it is the first
mechanization, in any proof assistant, of non-additive probability
over intuitionistic logic.

\begin{enumerate}
\item \textbf{The static hinge} (Section~\ref{sec:hinge}).  The
  classical complement rule holds for \emph{every} valuation on
  $\Omega$ iff excluded middle holds in $\Omega$.  The hard direction
  manufactures, at any failure of excluded middle, a valuation with
  positive slack from a prime ideal separation.  All of probability's
  classicality is thus located in one axiom, Van Horn's negation
  axiom R3 in the Cox tradition~\cite{vanhorn2003}.

\item \textbf{Representation theorems}
  (Section~\ref{sec:representation}).  On a finite frame, every
  valuation \emph{is} a classical probability mass function on the
  points, read through an interior operator.  It is literally a
  \textsf{mathlib} \texttt{PMF}, and equivalently a finite mixture of
  the frame's points.  On infinite frames this fails, by an explicit
  counterexample.  Scott continuity restores it, the counterexample
  is itself a proved failure of Scott continuity, and in general
  every valuation splits unconditionally into an atomic part (always
  a sub-probability) and a diffuse remainder, a constructive
  Lebesgue-style decomposition.

\item \textbf{The Dempster--Shafer bridge, completed}
  (Section~\ref{sec:ds}).  Inclusion--exclusion is constructively
  innocent: it holds \emph{with equality} for the intuitionistic
  join, on any frame.  Read through the double-negation nucleus, it
  yields the full $\infty$-monotone tower, so a valuation restricted
  to its Booleanization is a bona fide belief function.  Additive
  probability does not become a belief function by losing an axiom,
  but by being observed through $\neg\neg$.

\item \textbf{The dynamic hinge} (Section~\ref{sec:ds}).  The two
  classically indistinguishable update rules of Dempster--Shafer
  theory, Dempster's rule of conditioning and Bayesian (geometric)
  conditioning, are separated constructively by a single number.
  For evidence $b$ with $v(b) \neq 0$ they agree at every proposition
  iff $\slack(b) = 0$.  Classicality of the \emph{dynamics} is one
  instance of excluded middle, at the evidence.

\item \textbf{The Cox program, dissected} (Section~\ref{sec:cox} and
  Appendix~\ref{app:aczel}).  The product-rule half of Cox's theorem
  is logic-free, pure analysis, and we build its analytic core, an
  Acz\'el/H\"older-style associativity representation absent from
  \textsf{mathlib}, from scratch: a normalized, strictly monotone,
  additive generator constructed as a dyadic limit
  ($\approx$1{,}100 lines).  The sum-rule half is logically
  irreducible.  We exhibit a five-element frame on which a monotone
  normalized assignment is additive on disjoint joins yet not
  modular, and prove a disjunction's value is not any function of
  its disjuncts' values.

\item \textbf{Two grounding models} (Section~\ref{sec:bridges}).
  G\"odel--McKinsey--Tarski, quantitatively: any classical
  probability measure, read on the opens of a space, is a valuation,
  and slack is boundary mass.  And a computability guard: on the
  Sierpi\'nski frame of one semidecidable observation, the
  slack-free classical readout of ``machine $c$ halts'' is
  uncomputable by reduction to the halting problem, so any axiom
  forcing classical collapse demands uncomputable epistemic states.

\item \textbf{Product rigidity} (Section~\ref{sec:conclusion}).  On
  the chain product $\mathrm{Low}(\mathbb{N} \times \mathbb{N})$,
  every lower set is a finite union of rectangles, so two valuations
  agreeing on rectangles agree outright and the product valuation,
  whenever it exists, is unique with no continuity assumed.  This
  bears directly on whether a probability \emph{monad} exists for
  these valuations, a question whose hardness we delimit rather than
  resolve.
\end{enumerate}

\noindent
Section~\ref{sec:engineering} reports the formalization design and
the \textsf{mathlib}-upstreaming surface.
Section~\ref{sec:related} maps the relation to prior formalizations
(ALEA, HoTT valuations, Markov categories) and to the informal
literature.  Section~\ref{sec:conclusion} returns to the opening
theme, with the monad question stated precisely and the same
questions posed over \emph{linear} logic, where the event algebra is
a quantale and ``probability of a resource'' awaits its Gleason, as
the natural continuation of the program.
