%% =====================================================================
%% main.tex --- The Complement Rule Is Excluded Middle:
%%              Machine-Checked Probability over Intuitionistic Logic
%% Draft for CPP 2027 submission (acmart sigplan).
%%
%% Build: make          (pdflatex + bibtex + 2x pdflatex)
%% =====================================================================

\documentclass[sigplan,10pt,review,anonymous]{acmart}

\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{listingsutf8}
\usepackage{microtype}

% --- Lean 4 listing setup ---------------------------------------------
\lstdefinelanguage{Lean}{
  morekeywords={theorem,lemma,def,noncomputable,structure,instance,
                variable,import,namespace,end,open,scoped,where,
                fun,match,with,if,then,else,let,in,by,do,
                Prop,Type,Sort,deriving,inductive,abbrev,example,
                forall,exists,class,extends},
  sensitive=true,
  morecomment=[s]{/-}{-/},
  morecomment=[l]{--},
  literate=
    {->}{{$\rightarrow$}}{2}
    {→}{{$\rightarrow$}}{1}
    {↔}{{$\leftrightarrow$}}{1}
    {↦}{{$\mapsto$}}{1}
    {⟺}{{$\iff$}}{2}
    {∀}{{$\forall$}}{1}
    {∃}{{$\exists$}}{1}
    {∧}{{$\wedge$}}{1}
    {∨}{{$\vee$}}{1}
    {¬}{{$\neg$}}{1}
    {≤}{{$\le$}}{1}
    {≥}{{$\ge$}}{1}
    {≠}{{$\ne$}}{1}
    {∈}{{$\in$}}{1}
    {∉}{{$\notin$}}{1}
    {⊤}{{$\top$}}{1}
    {⊥}{{$\bot$}}{1}
    {⊔}{{$\sqcup$}}{1}
    {⊓}{{$\sqcap$}}{1}
    {⋁}{{$\bigvee$}}{1}
    {⨆}{{$\bigsqcup$}}{1}
    {∑}{{$\sum$}}{1}
    {∞}{{$\infty$}}{1}
    {ℝ}{{$\mathbb{R}$}}{1}
    {ℕ}{{$\mathbb{N}$}}{1}
    {ℤ}{{$\mathbb{Z}$}}{1}
    {Ω}{{$\Omega$}}{1}
    {δ}{{$\delta$}}{1}
    {μ}{{$\mu$}}{1}
    {ρ}{{$\rho$}}{1}
    {ι}{{$\iota$}}{1}
    {σ}{{$\sigma$}}{1}
    {∂}{{$\partial$}}{1}
    {×}{{$\times$}}{1}
    {⁻¹}{{${}^{-1}$}}{2}
    {ᶜ}{{${}^{\mathsf{c}}$}}{1}
    {ᶜᶜ}{{${}^{\mathsf{cc}}$}}{2}
    {⟨}{{$\langle$}}{1}
    {⟩}{{$\rangle$}}{1}
    {∘}{{$\circ$}}{1}
    {≡}{{$\equiv$}}{1}
    {∅}{{$\emptyset$}}{1}
    {□}{{$\Box$}}{1}
    {▷}{{$\triangleright$}}{1}
}

\lstset{
  language=Lean,
  inputencoding=utf8/latin1,
  extendedchars=true,
  basicstyle=\ttfamily\small,
  keywordstyle=\bfseries,
  commentstyle=\itshape,
  showstringspaces=false,
  breaklines=true,
  xleftmargin=1em,
  aboveskip=0.6em,
  belowskip=0.6em,
}

\acmConference[CPP '27]{Certified Programs and Proofs}{January 2027}{(submitted)}
\acmYear{2027}
\acmISBN{}
\acmDOI{}
\citestyle{acmauthoryear}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}
\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}
\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\emergencystretch=3em

% Shorthands
\newcommand{\ENNReal}{\ensuremath{[0,\infty]}}
\newcommand{\slack}{\mathrm{slack}}
\newcommand{\Bel}{\mathrm{Bel}}
\newcommand{\Pl}{\mathrm{Pl}}

\begin{document}

\title{The Complement Rule Is Excluded Middle:\\
  Machine-Checked Probability over Intuitionistic Logic}

\author{Anonymous Author(s)}
\affiliation{\institution{Anonymous Institution}\city{}\country{}}

\begin{abstract}
Kolmogorov's axioms fix not only an arithmetic of chance but a
\emph{logic} of events: a Boolean algebra.  There is no intrinsic
reason for that choice.  Different situations call for different
logics, and quantum mechanics long ago moved probability onto
non-Boolean event structures.  We mechanize, in Lean~4 over
\textsf{mathlib}, what probability becomes when the event logic is
\emph{intuitionistic}: valuations on frames (complete Heyting
algebras), the logic of verifiable propositions.  The development is
about 3{,}950 lines, \texttt{sorry}-free, and its landmark theorems
depend only on the three standard axioms.

Three families of machine-checked results emerge.  \emph{Hinges:}
the classical complement rule $v(\neg A) = 1 - v(A)$ holds for every
valuation exactly when excluded middle holds in the algebra, so all
of probability's classicality is one axiom.  Dynamically, Dempster's
rule of conditioning agrees with Bayesian conditioning exactly when
the evidence satisfies excluded middle.  \emph{Representation:} on
finite frames every valuation is a classical probability mass
function on points seen through an interior operator.  On infinite
frames this fails, Scott continuity restores it, and in general
there is an unconditional atomic/diffuse decomposition.  A valuation
is a Dempster--Shafer belief function on its Booleanization, totally
monotone, via an inclusion--exclusion identity that holds \emph{with
equality} for the intuitionistic join.  \emph{Guards:} a
Sierpi\'nski-space model built from a genuine measure shows the
theory's ``slack'' is boundary mass, and a reduction to the halting
problem shows the slack-free classical readout of a semidecidable
event is uncomputable: slack is the price of a computable epistemic
state.  Along the way the proof assistant caught an unsatisfiable
statement of the central Cox-style theorem and forced the
identification of which axiom is irreducible.  The analytic core
(Acz\'el's associativity theorem, H\"older-style) is built from
scratch as a dyadic-limit construction absent from \textsf{mathlib}.
\end{abstract}

\begin{CCSXML}
<ccs2012>
<concept>
<concept_id>10003752.10003790.10003792</concept_id>
<concept_desc>Theory of computation~Proof theory</concept_desc>
<concept_significance>500</concept_significance>
</concept>
<concept>
<concept_id>10003752.10003790.10011740</concept_id>
<concept_desc>Theory of computation~Type theory</concept_desc>
<concept_significance>300</concept_significance>
</concept>
<concept>
<concept_id>10002950.10003648</concept_id>
<concept_desc>Mathematics of computing~Probability and statistics</concept_desc>
<concept_significance>500</concept_significance>
</concept>
</ccs2012>
\end{CCSXML}
\ccsdesc[500]{Theory of computation~Proof theory}
\ccsdesc[500]{Mathematics of computing~Probability and statistics}

\keywords{intuitionistic logic, probability, Cox's theorem, belief
functions, locales, Lean, mathlib, formalized mathematics}

\maketitle

\input{intro}
\input{sec-valuation}
\input{sec-hinge}
\input{sec-representation}
\input{sec-ds}
\input{sec-cox}
\input{sec-bridges}
\input{sec-engineering}
\input{sec-related}
\input{sec-conclusion}

\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\appendix
\input{app-aczel}

\end{document}
