\section{Related Work}
\label{sec:related}

\paragraph{Informal precedents.}
That probability might be paired with intuitionistic logic is an old
if scattered thought.  Weatherson develops intuitionistic probability
axiomatically and defends it with Dutch-book
arguments~\cite{weatherson2003}.  Paris treats non-classical
uncertain reasoning broadly~\cite{paris1994}.  The identification
$\Bel = P(\Box\,\cdot)$ connecting belief functions to modal logic
is classical~\cite{ruspini1986,dempster1967,shafer1976}.  On the
localic side, valuations on frames are standard objects of
constructive measure theory, with Vickers's work and the
Coquand--Spitters development as canonical
references~\cite{vickers2008,coquand-spitters2009}, normally with
Scott continuity built in and with additive, Riesz-flavored goals.
What we take from Section~\ref{sec:representation} is that Scott
continuity is a substantive assumption (it decides
point-representability on the frames we examine), so our base
structure omits it.  The Cox line, through Cox, Jaynes, Halpern's
counterexample, and the Van Horn and Arnborg--Sj\"odin
repairs~\cite{cox1946,jaynes2003,halpern1999,vanhorn2003,
arnborg-sjodin2000}, is classical throughout.  Van Horn himself
observes that dropping R3 yields a two-number theory, which is
exactly the belief/plausibility pair our slack separates.  To our
knowledge, no prior work, informal or formal, states either hinge
theorem: complement rule $\iff$ excluded middle
(Theorem~\ref{thm:hinge}), or Dempster $=$ Bayes $\iff$ zero slack
at the evidence (Theorem~\ref{thm:cond-hinge}).

\paragraph{Prior formalizations.}
Three bodies of mechanized work are adjacent, and none overlaps.
ALEA~\cite{audebaud-paulin2009} formalizes a discrete distribution
monad in Coq for randomized programs: additive probability, no
event logic in our sense.  Recent work in cubical Agda formalizes
synthetic probability via Markov categories~\cite{sargsyan2023},
where again the objects are additive distributions.  Closest is the
HoTT/Coq line of Faissole and Spitters on \emph{valuations} over
lower reals~\cite{faissole-spitters2017}, done constructively with
Scott continuity and aimed at integration (Riesz) and a Giry-style
valuation \emph{monad}.  Their valuations are the additive,
continuous cousins of ours, and their monad is the Scott-continuous
instance of the question examined in
Section~\ref{sec:conclusion}.  The delta of this paper is not the
word ``valuation'' but the development on top of it: the
non-additive complement regime, slack and its decomposition, both
hinge theorems, the DS bridge with total monotonicity and the two
conditionings, the representation/mixture/Scott-atomic suite, the
corrected Cox theorem with its irreducibility counterexamples, the
Acz\'el construction, product rigidity, and the computability guard.
We claim first mechanization for that development, not for
valuations on locales as such.

\paragraph{Non-Boolean probability elsewhere.}
Quantum probability on orthomodular lattices, with Gleason's
theorem~\cite{birkhoff-vonneumann1936,gleason1957}, is the
long-standing precedent for re-basing probability on a non-Boolean
event logic.  Free probability, fuzzy measures, and imprecise
probability~\cite{walley1991} vary other parameters (independence,
additivity, single-valuedness).  Our variation is specifically the
\emph{logic}, holding the Cox-style methodology fixed.  Within DS
theory, the two conditionings and their inequivalence are
classical~\cite{shafer1976,fagin-halpern1991}.  What is new in
Theorem~\ref{thm:cond-hinge} is the exact logical characterization
of their agreement, and its mechanization.
