\section{Conclusion: Probability with the Logic as a Parameter}
\label{sec:conclusion}

We set out from the observation that Kolmogorov's axioms fix a logic
before they fix an arithmetic, and that nothing in the concept of
graded belief privileges the Boolean choice.  The machine-checked
results bear the framing out with a precision informal treatments
have not reached.  The entire classical surplus is one axiom, whose
statical face is the complement rule (Theorem~\ref{thm:hinge}) and
whose dynamical face is the agreement of Dempster and Bayesian
conditioning (Theorem~\ref{thm:cond-hinge}).  Everything else
survives the intuitionistic substrate intact: conditioning, Bayes
symmetry, disjoint additivity, total probability over genuine
partitions, and inclusion--exclusion itself
(Theorem~\ref{thm:incl-excl}).  What the weaker logic adds is
structure, not damage: slack with its two-gap decomposition, the
atomic/diffuse divide, the belief-function reading through
$\neg\neg$, and a computability-theoretic reason the additions are
not optional.

\paragraph{Open problems.}
Three are mathematical.  First, representing the diffuse part on
non-spatial frames (the localic Riesz direction).  Second, closing
the analytic residue of the Acz\'el development, namely continuity
of the generator and the off-cone extension (the cone results and
the log-sum-exp archetype are done).  Third, a genuine uniqueness
theorem: deriving modularity from a more primitive desideratum,
knowing (Theorem~\ref{thm:no-functional}) that it cannot come from
disjunction data alone, with the mixture characterization
(Theorem~\ref{thm:mixture}) as the current best justification.  A
fourth is foundational hygiene at scale: a machine-checked
Dutch-book theorem in the style of~\cite{weatherson2003}, which to
our knowledge no proof assistant has for \emph{any} logic.

\paragraph{The functor is literal, the monad is delimited.}
Within a fixed doctrine the connection to the categorical picture is
concrete.  Pushforward along frame homomorphisms preserves all four
valuation axioms, making $\mathrm{Val}(-)$ a functor on locales, the
localic face of the functor part of the Giry
monad~\cite{giry1982}.  This is a routine verification that we state
but have not mechanized.  The point valuations $\delta_p$ of
Section~\ref{sec:representation} are the evident unit (these
\emph{are} formalized), and Theorem~\ref{thm:finite-rep} shows that
on finite frames $\mathrm{Val}$ agrees, on objects, with the
classical finite-distribution monad on points.  The
\emph{multiplication} is where the substance is.  Flattening a
valuation of valuations is integration, hence product valuations and
Fubini, and by Kock's identification of commutative monads with
Fubini theorems~\cite{kock2012}, whether
statistics-as-kernel-composition transports to a weaker logic is
precisely the question whether that logic's valuation functor is a
commutative monad.  Two machine-checked results now delimit the
multiplication from both sides.

On the chain product the question of \emph{which} product valuation
to take does not arise.  The frame
$\mathrm{Low}(\mathbb{N} \times \mathbb{N})$ is \emph{rigid}.
Every lower set is a \emph{finite} union of rectangles: beyond a
stabilization height every occupied column is full, so the full
columns and finitely many rows account for everything
(\texttt{exists\_rect\_decomposition}).
The rigidity mechanism is general: by the inclusion--exclusion
equality, two valuations agreeing on any $\sqcap$-closed family
agree on all finite joins of its members
(\texttt{eq\_on\_finsetSup\_of\_eq\_on}), and the chain product is
the instance where that family is the rectangles.  So the product
valuation there, whenever it exists, is unique, with no continuity
assumed (\texttt{Valuation.product\_unique}).  We had expected the
opposite, diffuse mass free to choose among the infinities of the
product, and our candidate counterexamples all died on the same
rock: lower sets of chain products detect only finitely much of the
boundary.

They were right in spirit and wrong in venue.  On the powerset frame
$\mathrm{Set}(\mathbb{N} \times \mathbb{N})$, whose elements are
far from finite unions of rectangles, non-uniqueness is a theorem.
The indicator of an ultrafilter is a valuation
(\texttt{ultrafilterValuation}), and any ultrafilter on a product
splits rectangles through its marginal pushforwards
(\texttt{rect\_mem\_iff}).  Taking a free ultrafilter
$\mathcal{U}$ on $\mathbb{N}$, the two iteration orders of its
tensor with itself and its diagonal pushforward all carry identical
values on \emph{every} rectangle, yet they differ on the upper
triangle $\{(m, m') \mid m < m'\}$
(\texttt{product\_valuation\_not\_unique}).  In particular the two
orders of iterated integration of the same marginal data disagree:
\emph{Fubini fails}, concretely
(\texttt{fubini\_fails\_for\_valuations}).  Compare the Fubini
pathologies of non-additive
integration~\cite{denneberg1994,ghirardato1997}.  By Kock's
correspondence, no canonical commutative multiplication exists for
unrestricted valuations.  On the powerset frame these valuations are
finitely additive probability charges in the sense
of~\cite{bhaskara-rao1983}, so this half of the dichotomy is not
about intuitionism at all: the freedom lives already in the
classical finitely additive world, and dropping continuity is what
lets it in.  Anyone who wants a monad must therefore restrict, and
the dichotomy says along which axis: the frames, or the
valuations.

\paragraph{The hard problem.}
The question \emph{upstream} of the multiplication deserves to have
its hardness named rather than implied, and what follows is our
assessment rather than a theorem.  A monad needs a category for
$\mathrm{Val}$ to be an endofunctor on.  The space of valuations
must itself be an object, a ``space'' whose opens one can speak of.
For Scott-continuous valuations that space is a locale and monad
structure is available, constructively, through the probabilistic
powerdomain
line~\cite{jones-plotkin1989,vickers2008,coquand-spitters2009}.
Yet \emph{even there}, exhibiting a cartesian closed category
closed under the construction is the Jung--Tix problem, open since
1998~\cite{jung-tix1998}.  Dropping continuity makes the difficulty
prior rather than merely worse: the standard presentations of the
valuation space use continuity essentially.  One route to a
well-posed question does exist.  The localic treatments present the
space of continuous valuations as the classifying locale of a
propositional geometric theory~\cite{vickers1989,
coquand-spitters2009}, with generators asserting $q < v(a)$ for
rational $q$ and axioms for monotonicity, normalization, modularity
(expressible with rational witnesses and countable disjunctions),
and continuity.  Deleting the continuity axiom leaves a geometric
theory, so a classifying locale still exists on general grounds, and
classically its points are exactly the valuations of this paper.  We
have verified none of the structure this candidate would need.  The
unit looks unproblematic, but the multiplication requires
integrating the evaluation maps against a merely modular,
discontinuous valuation, and there the sum of two lower-real-valued
maps is an infinite join, exactly where the diffuse part of
Theorem~\ref{thm:obstruction} leaks.  We offer the formulation as a
target, not a result.

The strategy with the best modern track record is to carve a
fragment rather than to conquer the whole.  On the measure-theoretic
side, the failure of cartesian closure for measurable spaces was
answered by quasi-Borel spaces~\cite{heunen2017qbs}.  On the
domain-theoretic side, commutativity of the full valuations monad on
\textsf{DCPO} is open, and the response is a hierarchy of commutative
\emph{submonads} generated from simple
valuations~\cite{jia2021commutative}.  Our decomposition theorem
names the corresponding fragment here: the atomic part.  As a first
machine-checked step in that direction we prove that the
countably-presented fragment composes with no space structure at
all.  A countable mixture of valuations is a valuation
(\texttt{mixCountable}), a mixture of mixtures flattens to a mixture
with product weights, which is discrete Chapman--Kolmogorov
(\texttt{mixCountable\_mixCountable}), and the one-point mixture is
the identity (\texttt{mixCountable\_unique}), all by the
unconditional Fubini of $\ENNReal$ sums over sigma types.  And in
exact contrast with the failure above, the fragment is
\emph{commutative}: iterated countable mixing is order-independent
(\texttt{mixCountable\_comm}).  Commutative exactly where countable.
This is associative, unital, commutative kernel composition for
countably-atomic data.  It is not yet a monad: the presentations are
not canonical, we do not quotient by ``presents the same
valuation,'' and the diffuse part is exactly what escapes every
countable presentation.  The ledger, then.  The unit, the finite
level, countable kernel composition, and its commutativity are
proved.  Rigidity on chain products and non-uniqueness on the
powerset frame are proved, so the general multiplication is settled:
it is not canonically determined.  What remains is the ambient
category, a formulation problem before it is anything else.

\paragraph{Future work on the monad.}
Two continuations are concrete.  First, quotient the countable
presentations by ``presents the same valuation'' and package the
result as a Kleisli category of countably-atomic kernels, checking
that mixing descends to the quotient.  That would upgrade
composition-on-data to an honest categorical object, and we expect
it to be a development of its own.  Second, mechanize the
classifying-locale formulation sketched above, which requires
infrastructure \textsf{mathlib} currently lacks entirely:
classifying locales of propositional geometric theories.  Both
continuations inherit the boundary drawn here.  Whatever structure
exists must live on a fragment, because the unrestricted
multiplication is now provably not canonical.

Across doctrines, by contrast, we make no functoriality claim.
There is no ambient category in which ``logic $\mapsto$ its
probability theory'' is a functor, and each column requires
re-deriving what a state \emph{is} from the desiderata.  What is
uniform across columns is the \emph{method}, and this paper's
execution of it is the Heyting column.

\paragraph{The program.}
If the logic is a parameter, this paper computed one new value of
it.  The table wants filling in.  Linear logic is the most inviting
column for this community.  Its event algebras are quantales, its
conjunction is resource-sensitive, and a ``valuation'' there should
grade the \emph{consumption} of evidence.  The analogue of Gleason's
theorem or of our representation theorems (which states are
compatible with the algebra?) is wide open, as is the right analogue
of slack for the failure of contraction.  The probabilistic
semantics of linear logic that already exist, probabilistic
coherence spaces~\cite{danos-ehrhard2011}, start from the additive
side.  A Cox-style derivation over a quantale would start from the
logic.  The methodology of this paper, to state the classical
desiderata, delete the axioms the logic no longer licenses, and let
a proof assistant referee what survives, what splits, and what
becomes unsatisfiable, is, we suggest, the right instrument for that
program.  Each cell of the table is exactly the kind of delicate,
counterexample-rich boundary work that mechanization is for.
