\section{Where Classicality Lives: the R3 Hinge}
\label{sec:hinge}

Van Horn's repaired Cox theorem~\cite{vanhorn2003} axiomatizes
plausibility over classical propositions.  Its negation axiom, called
\emph{R3}, posits that the plausibility of $\neg A$ is a fixed
non-increasing function $S$ of the plausibility of $A$.  Consistency
forces $S \circ S = \mathrm{id}$, an involution, and thence the
complement rule $p \mapsto 1 - p$.  An involutive negation \emph{is}
double-negation elimination, so R3 is where classical logic enters
every known Cox-style derivation.  We isolate the exact content of
R3's conclusion as a predicate on valuations and prove it equivalent
to excluded middle.

\begin{lstlisting}
def HasClassicalNegation (v : Valuation Ω) : Prop :=
  ∀ a, v a + v aᶜ = 1
\end{lstlisting}

\begin{theorem}[the R3 hinge]
\label{thm:hinge}
For a frame $\Omega$, every valuation on $\Omega$ has classical
negation \emph{iff} $a \sqcup a^{\mathsf{c}} = \top$ for every
$a \in \Omega$.
\end{theorem}

The forward direction (\texttt{hasClassicalNegation\_of\_em}) is a
calculation.  The content is the converse
(\texttt{em\_of\_forall\_hasClassicalNegation}): wherever excluded
middle fails, a valuation with genuine slack must be
\emph{manufactured}.  Suppose $a \sqcup a^{\mathsf{c}} \neq \top$.
The prime ideal separation theorem
(\textsf{mathlib}'s \texttt{DistribLattice.prime\_ideal\_of\_%
disjoint\_filter\_ideal}) yields a prime ideal $J$ containing
$a \sqcup a^{\mathsf{c}}$ but not $\top$.  The indicator of $J$'s
complement,
\[
  v(x) = \begin{cases} 0 & x \in J \\ 1 & x \notin J, \end{cases}
\]
is a valuation.  Modularity in the only nontrivial case ($x, y
\notin J$) is \emph{exactly} primeness, $x \sqcap y \in J
\Rightarrow x \in J \vee y \in J$.  And this valuation assigns
$v(a) + v(a^{\mathsf{c}}) = 0 \neq 1$.  So the complement rule is not
one convention among many.  Quantified over all valuations, it is a
statement \emph{about the algebra}, and the statement is excluded
middle.

\paragraph{Sharp valuations are the points of the locale.}
The construction above is not ad hoc.  It is one half of an exact
correspondence.  Call $v$ \emph{sharp} if it is $\{0,1\}$-valued, a
state of full certainty.

\begin{theorem}[\texttt{sharp\_iff\_point}]
$v$ is sharp iff $v$ is the complement-indicator of a prime ideal.
The maps are mutually inverse: the ideal recovered from $v$ is its
zero-set, prime exactly because $v$ is sharp.
\end{theorem}

\noindent
In locale-theoretic terms, the certainty limit of the probability
theory recovers the \emph{points} of the locale.  (These are the
finitely prime points.  The completely prime points would correspond
to sharp valuations that are also Scott-continuous, a refinement we
note but do not formalize.)  The hinge and the correspondence
together are the paper's static answer to ``where does classicality
live'': in one axiom, whose semantic shadow is point-hood.
