\section{The Dempster--Shafer Reading, Completed}
\label{sec:ds}

Sections~\ref{sec:valuation}--\ref{sec:representation} keep meeting
the \emph{shape} of Dempster--Shafer (DS)
theory~\cite{dempster1967,shafer1976}: sub-additive complements, a
belief/plausibility interval, mass functions.  This section makes the
identification exact in both statics and dynamics, and locates, each
time, the precise point where excluded middle is the difference.

\subsection{Statics: total monotonicity via constructive
inclusion--exclusion}

The $\neg\neg$-stable (\emph{regular}) elements of a frame form a
Boolean algebra with meet $\sqcap$ and join
$(a \sqcup b)^{\mathsf{cc}}$ (Glivenko).  Call it the
\emph{Booleanization}.  A DS belief function is a capacity on a
Boolean algebra that is $n$-monotone for every $n$:
$\Bel(\bigvee_i A_i) \ge \sum_{\emptyset \neq T}
(-1)^{|T|+1} \Bel(\bigwedge_{i \in T} A_i)$.  The $2$-monotone case
is a one-liner from modularity
(\texttt{two\_monotone}).  The full tower follows from a fact that is
interesting in its own right:

\begin{theorem}[inclusion--exclusion,
  \texttt{inclusion\_exclusion}]
\label{thm:incl-excl}
For every valuation on every frame and every finite family
$a : \iota \to \Omega$, inclusion--exclusion holds \emph{with
equality} for the frame join:
\[
  v\Bigl(\,\bigsqcup_{i \in s} a_i\Bigr)
  \;=\; \sum_{\emptyset \neq T \subseteq s} (-1)^{|T|+1}\,
        v\Bigl(\,\bigsqcap_{i \in T} a_i\Bigr).
\]
(Stated in Lean additively, with the odd-cardinality half of the
signed sum set against the even-cardinality half, so that no
$\ENNReal$ subtraction occurs.  The proof is induction with
modularity and frame distributivity.)
\end{theorem}

Reading the same signed sum against the Booleanization join turns
the equality into the full belief-function tower:

\begin{theorem}[total monotonicity, \texttt{infty\_monotone}]
\label{thm:infty}
For every valuation, finite family, and $n$:
\[
  \sum_{\emptyset \neq T} (-1)^{|T|+1}
  v\Bigl(\bigsqcap_{T} a_i\Bigr)
  \;\le\;
  v\Bigl(\bigl(\bigsqcup_i a_i\bigr)^{\mathsf{cc}}\Bigr).
\]
Read on the Booleanization, where $(\cdot)^{\mathsf{cc}} \circ
\sqcup$ is the join and meets of regulars are regular, this is
verbatim the $\infty$-monotone tower: $v$ restricted to its
Booleanization is a bona fide belief function.
\end{theorem}

The proof of Theorem~\ref{thm:infty} from
Theorem~\ref{thm:incl-excl} is one line: monotonicity along
$a \le a^{\mathsf{cc}}$.  The conceptual point runs opposite to the
usual telling.  Inclusion--exclusion does \emph{not} fail
constructively.  It is exact for the intuitionistic disjunction, and
total monotonicity is what that equality looks like through the
double-negation nucleus: rounding a join up to its regular hull can
only add value, and the defect is exactly the double-negation gap of
Section~\ref{sec:valuation}.  Additive probability does not
\emph{become} a belief function by losing an axiom.  It becomes one
by being \emph{read through} $\neg\neg$.  With
$\Pl(a) = 1 - v(a^{\mathsf{c}})$, the belief--plausibility interval
has width exactly the slack
(\texttt{plausibility\_sub\_self}).  The DS ``ignorance'' is
boundary measure, not psychology.

\subsection{Dynamics: Dempster against Bayes, separated by one
number}

DS theory carries two update rules that coincide classically.
\emph{Geometric} conditioning renormalizes belief, and under the DS
dictionary it \emph{is} the localic posterior
$v(a \sqcap b)/v(b)$ of Section~\ref{sec:valuation}.
\emph{Dempster's rule of conditioning}~\cite{shafer1976} discards
the mass incompatible with the evidence and renormalizes by its
plausibility:
\[
  v(a \,\|\, b) \;=\;
  \frac{v(a \sqcup b^{\mathsf{c}}) - v(b^{\mathsf{c}})}
       {1 - v(b^{\mathsf{c}})}.
\]
Both are again valuations (\texttt{dempsterCond}, whose modularity
is the frame distributivity
$(a \sqcup b^{\mathsf{c}}) \sqcap (a' \sqcup b^{\mathsf{c}})
= (a \sqcap a') \sqcup b^{\mathsf{c}}$).  Their disagreement is
governed by a single structural quantity, the \emph{excluded-middle
gap} $\mathrm{emGap}(a,b) = v(a) - v(a \sqcap (b \sqcup
b^{\mathsf{c}}))$, the mass of $a$ stranded outside $b$'s instance
of excluded middle:

\begin{lemma}[\texttt{sup\_compl\_decomp}]
$v(a \sqcup b^{\mathsf{c}}) = v(b^{\mathsf{c}}) + v(a \sqcap b) +
\mathrm{emGap}(a,b)$.  Dempster's numerator exceeds Bayes's by
exactly the gap.
\end{lemma}

The gap, and with it the disagreement of the two update rules, is
controlled by one number, the slack at the evidence:

\begin{theorem}[the conditioning hinge,
  \texttt{dempsterCond\_eq\_condVal\_iff\_slack}]
\label{thm:cond-hinge}
For $v(b) \neq 0$: Dempster conditioning on $b$ agrees with
Bayesian/geometric conditioning on $b$ at \emph{every} proposition
iff $\slack(b) = 0$.
\end{theorem}

Zero slack at the evidence kills the gap for every $a$ and makes the
two normalizers coincide, since $1 - v(b^{\mathsf{c}}) = v(b)$.
Conversely, testing at $a := b$, geometric conditioning gives
certainty $1$ while Dempster gives $v(b)/(1 - v(b^{\mathsf{c}}))$,
which is $1$ only if the slack vanishes.
Theorem~\ref{thm:cond-hinge} is the dynamic companion of
Theorem~\ref{thm:hinge}.  The static hinge locates classicality in
the complement rule, and the dynamic hinge locates it in the
agreement of the two DS update rules: one instance of excluded
middle, at the evidence.  In the measure model, Dempster and Bayes
disagree exactly when the evidence has a charged boundary.
