\section{The Cox Program, Dissected}
\label{sec:cox}

Cox's theorem~\cite{cox1946,jaynes2003} derives the probability
calculus from desiderata on plausibility.  After Halpern's
counterexample~\cite{halpern1999}, its repaired
forms~\cite{vanhorn2003,arnborg-sjodin2000} fix the calculus up to
monotone regraduation.  Every known derivation is classical, and
Section~\ref{sec:hinge} showed why: the negation axiom R3 \emph{is}
excluded middle.  What survives of the program when R3 is dropped?
Mechanization forced a sharper answer than we had conjectured, in
three parts.  The product rule is logic-free.  The sum rule is
irreducible.  And the naive constructive statement is unsatisfiable.

\subsection{A Cox structure without negation}

\begin{lstlisting}
structure CoxModel (Ω : Type*) [Order.Frame Ω] where
  pl : Ω → Ω → ℝ            -- pl a b : plausibility of a given b
  mono_left : ∀ c, Monotone fun a => pl a c
  pl_bot : ∀ c, pl ⊥ c = 0
  pl_top : ∀ c, pl ⊤ c = 1
  F : ℝ → ℝ → ℝ             -- conjunction functional
  conj : ∀ a b c, pl (a ⊓ b) c = F (pl a (b ⊓ c)) (pl b c)
  F_assoc : ∀ x y z, F (F x y) z = F x (F y z)
  F_strictMono_left : ∀ y, 0 < y → StrictMono fun x => F x y
\end{lstlisting}

The deliberate omission is any field relating
$\mathrm{pl}(a^{\mathsf{c}}, b)$ to $\mathrm{pl}(a, b)$.  That
missing field is exactly R3.  Two details are themselves
checker-forced.  Strict monotonicity of $F$ must be restricted to
$0 < y$, because the boundary axioms force $F(x, 0) = 0$ for all
$x$, so the unrestricted axiom makes the structure \emph{uninhabited}
and every theorem about it vacuous.  We guard against this
permanently: \texttt{coxModelENNReal} is an explicit model on the
non-Boolean chain $\ENNReal$, built without classical logic (deciding
$a = \top$ there is a constructor check), so the axiom class is
inhabited by a genuinely non-classical instance.

\subsection{The sum rule is irreducible}

Classically, additivity is \emph{derived}: R3 turns disjunction into
negated conjunction via De~Morgan, and the product rule does the
rest.  Constructively that route is closed, since
$a \vee b \neq \neg(\neg a \wedge \neg b)$ in general.  And this is
not an artifact of one proof strategy:

\begin{theorem}[\texttt{modularity\_irreducible},
  \texttt{no\_disjunction\_functional}]
\label{thm:no-functional}
There is a five-element frame (the lower sets of the poset
$o < a,\, o < b$ with $a, b$ incomparable) and a monotone, normalized
assignment $q$ on it that is additive on disjoint joins yet not
modular.  Moreover no binary operation $S$ on values satisfies
$q(x \sqcup y) = S(q(x), q(y))$ for all $x, y$.  The pairs
$({\downarrow}a, {\downarrow}a)$ and $({\downarrow}a, {\downarrow}b)$
have identical value-pairs $(\tfrac12, \tfrac12)$ but joins valued
$\tfrac12$ and $1$.
\end{theorem}

\noindent
A disjunction's plausibility is not a function of its disjuncts'
plausibilities, unlike a conjunction's, which the \texttt{conj}
field reduces to values via conditioning.  So a constructive Cox
theorem \emph{must} posit the sum-rule half.  Modularity is the
posit, and Theorem~\ref{thm:mixture} is its independent
justification: modularity says exactly that the calculus is closed
under mixing certainties.

\subsection{The theorem, corrected and proved}

Our original target statement asked for a strictly monotone
regraduation $g : \mathbb{R} \to \ENNReal$ with $g(0) = 0$.  The
checker refused, for a reason no referee had caught on paper: strict
monotonicity forces $g(-1) < g(0) = 0$, impossible in $\ENNReal$.
The statement was not too hard.  It was \emph{unsatisfiable}, and
the repair (strictness only on $[0,1]$, where plausibilities live)
is exactly the domain hygiene that Halpern's critique of the
classical theorem turns on.  The corrected statement is a theorem:

\begin{theorem}[\texttt{constructive\_cox}]
\label{thm:cox}
Every Cox model whose unconditional plausibility is modular
regraduates to a valuation: there exist $v : \mathrm{Valuation}\;
\Omega$ and $g$ strictly monotone on $[0,1]$ with $g(0)=0$,
$g(1)=1$, and $v(a) = g(\mathrm{pl}(a, \top))$ for all $a$.
\end{theorem}

The regraduation for the sum-rule half is undemanding
($g = \texttt{ENNReal.ofReal}$).  The substance lives in the
product-rule half, showing $F$ regraduates to multiplication, which
is Acz\'el's associativity theorem~\cite{aczel1966}, a statement of
pure analysis whose type never mentions $\Omega$.  The
constructive/classical divide is therefore \emph{entirely} in the
sum rule.  \textsf{mathlib} has no Acz\'el theorem, no H\"older-style
embedding for ordered semigroups on intervals, and no Cauchy
functional equation under monotonicity (only under continuity), so
we build the analytic core from scratch ($\approx$1{,}100 lines): a
bundled \texttt{Scale} of Acz\'el hypotheses, a generator defined as
the dyadic limit $g(b) = \bigsqcup_n (\text{count of } 2^n\text{-th
roots under } b)/2^n$, and the H\"older-style conclusion, a
normalized, strictly monotone additive generator and hence a
strictly monotone multiplicative one, on the positive cone
(\texttt{exists\_mul\_generator}, \texttt{aczelStatement\_cone}).
The Archimedean property is \emph{derived} from continuity and
positivity rather than assumed, and the interval form of
divisibility is likewise a consequence of the intermediate value
theorem (\texttt{exists\_diag\_eq}), though the bundled
\texttt{Scale} takes divisibility as a hypothesis.
Appendix~\ref{app:aczel} gives the construction.  The residual
analytic gap, continuity of the generator and extension below the
unit, is stated honestly there and in
Section~\ref{sec:conclusion}, and is witnessed non-vacuously by the
log-sum-exp archetype, for which the global continuous generator is
$\exp$.
