\section{Representation: Points, Masses, Mixtures}
\label{sec:representation}

Is intuitionistic probability a new animal, or classical probability
seen at an angle?  The answer is exact: on finite frames the latter,
in general the former, with Scott continuity marking the boundary.

\subsection{Finite frames: every valuation is a PMF on points}

Every finite frame is the lattice of \emph{lower sets} of a finite
poset $P$, by Birkhoff duality.  We work with \texttt{LowerSet P}
directly and cite the duality rather than mechanize it.  Define the
\emph{mass} of a point as the discrete derivative of $v$:
\begin{lstlisting}
def Valuation.mass (v : Valuation (LowerSet P)) (p : P) :
    ℝ≥0∞ := v (LowerSet.Iic p) - v (LowerSet.Iio p)
\end{lstlisting}

\begin{theorem}[finite representation,
  \texttt{eq\_sum\_mass}, \texttt{sum\_mass}]
\label{thm:finite-rep}
For finite $P$ and any valuation $v$ on $\mathrm{Low}(P)$:
$v(U) = \sum_{p \in U} \mathrm{mass}(p)$ and
$\sum_{p} \mathrm{mass}(p) = 1$.
\end{theorem}

The proof peels a maximal point $p$ of $U$ at a time.  Modularity
applied to $(U \setminus {\uparrow}p,\; {\downarrow}p)$ shows the
increment is exactly $\mathrm{mass}(p)$, and a strong induction on
$|U|$ finishes.  The arithmetic is in $\ENNReal$ with truncated
subtraction, where the bookkeeping is real work.  Two packagings make
the classicality literal.  \texttt{Valuation.toPMF} exhibits
$\mathrm{mass}$ as a \textsf{mathlib} \texttt{PMF} on $P$.  And since
lower sets are the opens of the Alexandrov topology, where interior
is the identity, the theorem reads: every valuation on a finite
frame is a classical probability seen through the interior operator,
$v = \mu \circ \Box$.  This is the converse, in the finite case, of
the measure-theoretic bridge of Section~\ref{sec:bridges}.

Dually, with $\delta_p$ the sharp \emph{point-valuation}
$\delta_p(U) = [p \in U]$:

\begin{theorem}[mixture characterization,
  \texttt{eq\_mix\_deltaPoint}]
\label{thm:mixture}
On a finite frame, $v = \sum_p \mathrm{mass}(p)\cdot\delta_p$:
valuations are exactly the finite convex mixtures of
point-valuations.
\end{theorem}

\noindent
Modularity is thereby \emph{characterized} as closure under mixing
certainties.  This is the constructive answer to what the sum rule
axiomatizes, given that it cannot be derived
(Section~\ref{sec:cox}).

\subsection{The infinite boundary: enter Scott continuity}

Both halves of the finite theorem fail at the first infinite frame,
and the failure is informative.

\begin{theorem}[obstruction,
  \texttt{exists\_valuation\_not\_point\_representable}]
\label{thm:obstruction}
On $\mathrm{Low}(\mathbb{N})$, the indicator of $\top$ is a valuation
with $\mathrm{mass}(p) = 0$ at every point yet $v(\top) = 1$.  The
unit of mass escapes to the missing point at infinity.
\end{theorem}

\begin{theorem}[recovery, \texttt{eq\_tsum\_mass\_of\_scott},
  generalized by \texttt{isPurelyAtomic\_of\_scott}]
On $\mathrm{Low}(\mathbb{N})$, Scott continuity restores
point-representability: if $v(\top) = \bigsqcup_n v(\mathrm{Iic}\,n)$
then $v(\top) = \sum'_n \mathrm{mass}(n)$.  The counterexample of
Theorem~\ref{thm:obstruction} is itself a proved failure of Scott
continuity (\texttt{topIndicator\_not\_scott}).  More generally, on
any poset all of whose principal downsets are finite, Scott
continuity implies zero diffuse part.
\end{theorem}

\begin{theorem}[general decomposition, \texttt{tsum\_mass\_le}]
\label{thm:decomposition}
On $\mathrm{Low}(P)$ for \emph{arbitrary} $P$, with no finiteness and
no continuity assumption:
$\sum'_p \mathrm{mass}(p) \le v(\top)$.
\end{theorem}

\noindent
Every valuation thus splits unconditionally into an \emph{atomic}
part carried by points, always a sub-probability, and a
\emph{diffuse} remainder $v(\top) - \sum'_p \mathrm{mass}(p) \ge 0$.
This is a constructive Lebesgue-style decomposition.  The finite and
Scott cases are the purely atomic extreme, and the $\top$-indicator
is the purely diffuse one.  The proof of
Theorem~\ref{thm:decomposition} re-runs the maximal-point peel over
arbitrary finite subsets, bounded by the whole, and passes to the
supremum.  It uses neither excluded middle nor choice beyond
\textsf{mathlib}'s baseline.  Representing the diffuse part on
non-spatial frames, where, tellingly, the obstruction is \emph{too
few points}, the same phenomenon as undecidability in
Section~\ref{sec:bridges}, is the open frontier
(Section~\ref{sec:conclusion}).
