\section{Valuations on Frames}
\label{sec:valuation}

\subsection{The event logic}

A \emph{frame} (locale) is a complete lattice in which finite meets
distribute over arbitrary joins.  Equivalently, it is a complete
Heyting algebra.  Frames are the algebraic semantics of
intuitionistic logic and the point-free abstraction of topological
spaces.  The motivating example throughout is
$\Omega = \mathcal{O}(X)$, the opens of a space, with
$\sqcap, \sqcup$ as intersection and union and the pseudo-complement
$a^{\mathsf{c}}$ of an open given by the \emph{interior} of its set
complement.  In \textsf{mathlib} this is the typeclass
\texttt{Order.Frame}, and the Heyting operations come with the key
asymmetry already in place: $a \sqcap a^{\mathsf{c}} = \bot$ holds in
any Heyting algebra (\texttt{inf\_compl\_eq\_bot}), while
$a \sqcup a^{\mathsf{c}} = \top$ is available only in Boolean
algebras.  An element with $a \sqcup a^{\mathsf{c}} = \top$ is
\emph{complemented}: it satisfies its own instance of excluded
middle.

\subsection{The valuation structure}

\begin{definition}[\texttt{Valuation}]
\label{def:valuation}
A \emph{valuation} on a frame $\Omega$ is a function
$v : \Omega \to \ENNReal$ with
$v(\bot) = 0$, $v(\top) = 1$, monotonicity, and \emph{modularity}:
\[
  v(a) + v(b) \;=\; v(a \sqcup b) + v(a \sqcap b).
\]
\end{definition}

\begin{lstlisting}
structure Valuation (╬® : Type*) [Order.Frame ╬®] where
  toFun : ╬® ŌåÆ ŌäØŌēź0Ōł×
  map_bot' : toFun ŌŖź = 0
  map_top' : toFun ŌŖż = 1
  mono'    : Monotone toFun
  modular' : ŌłĆ a b, toFun a + toFun b
             = toFun (a ŌŖö b) + toFun (a ŌŖō b)
\end{lstlisting}

Values live in $\ENNReal$ (\texttt{ENNReal}, extended non-negative
reals) following \textsf{mathlib}'s measure-theoretic convention,
and normalization then confines them to $[0,1]$
(\texttt{Valuation.le\_one}).  The definition is Kolmogorov's minus
one thing: \emph{no axiom mentions complement}.  Classically that
omission is invisible, because the complement rule is derivable from
additivity and $a \vee \neg a = \top$.  Constructively it is the
whole story.

\subsection{Slack, and its two independent obstructions}

Modularity applied to the disjoint pair $(a, a^{\mathsf{c}})$
collapses to
$v(a) + v(a^{\mathsf{c}}) = v(a \sqcup a^{\mathsf{c}})$
(\texttt{add\_compl\_eq\_sup}): the sum measures $a$'s instance of
excluded middle.  Since that join need not be $\top$:

\begin{theorem}[\texttt{add\_compl\_le\_one}]
For every valuation and every $a$:
$v(a) + v(a^{\mathsf{c}}) \le 1$.
\end{theorem}

\noindent
This is the Dempster--Shafer sub-additivity inequality, here
\emph{derived} from the logic rather than postulated.  The deficit
\[
  \slack(a) \;=\; 1 - \bigl(v(a) + v(a^{\mathsf{c}})\bigr)
  \;=\; 1 - v(a \sqcup a^{\mathsf{c}})
\]
is the valuation of the undecided region.  It decomposes canonically.
The pair $(a^{\mathsf{cc}}, a^{\mathsf{c}})$, double negation
against negation, is always disjoint but is a partition only under
\emph{weak} excluded middle.  Running it through modularity splits
the slack into two logically independent gaps:

\begin{theorem}[\texttt{slack\_eq\_dnGap\_add\_deMorganGap}]
$\slack(a) = \mathrm{dnGap}(a) + \mathrm{deMorganGap}(a)$, where
$\mathrm{dnGap}(a) = v(a^{\mathsf{cc}}) - v(a)$ vanishes when $a$ is
$\neg\neg$-stable (regular), and
$\mathrm{deMorganGap}(a) = 1 - v(a^{\mathsf{cc}} \sqcup a^{\mathsf{c}})$
vanishes when weak excluded middle holds at $a$.
\end{theorem}

\noindent
The single Dempster--Shafer ``ignorance'' number is thus resolved
into a regularity defect plus a De~Morgan defect, each tracking a
named intermediate logic.

\subsection{What survives without excluded middle}

Three classical pillars survive intact, and one fails in a precisely
delimited way.  \emph{Disjoint additivity} is unconditional:
$a \sqcap b = \bot$ implies $v(a \sqcup b) = v(a) + v(b)$.
\emph{Conditioning} is native: for $v(b) \neq 0$ the posterior
$v(a \mid b) = v(a \sqcap b)/v(b)$ is again a valuation (the proof is
frame distributivity), and the product rule
$v(a \mid b) \cdot v(b) = v(a \sqcap b)$ and Bayes symmetry
$v(a \mid b)\,v(b) = v(b \mid a)\,v(a)$ hold
(\texttt{condVal\_mul}, \texttt{condVal\_symm}).  \emph{Total
probability} holds over any genuine partition, meaning
$a \sqcap a' = \bot$ and $a \sqcup a' = \top$:
$v(b) = v(a \sqcap b) + v(a' \sqcap b)$.  What fails is total
probability over $\{a, \neg a\}$.  That family does not tile
$\top$, and the conditional masses fall short of $v(b)$ by a
conditional slack (\texttt{cond\_add\_compl\_le}).  Bayesian updating
survives.  The assumption that $A$ and $\neg A$ exhaust the world
does not.

\paragraph{The classical fragment, a correction forced by the
checker.}
Where does honestly additive probability live inside $\Omega$?  The
natural guess is the $\neg\neg$-stable (\emph{regular}) elements,
which form a Boolean algebra by Glivenko's theorem.  The guess is
\emph{wrong}, and the failed proof said why: the valuation's
modularity governs the frame join $\sqcup$, while the regular
elements' Boolean join is the regularized
$(\cdot \sqcup \cdot)^{\mathsf{cc}}$.  A regular but uncomplemented
element (the open ray $(-\infty,0)$ in the opens of $\mathbb{R}$,
under an atomic measure at $0$) still carries slack.  The correct
classical fragment is the \emph{complemented} elements, where
$v(a) + v(a^{\mathsf{c}}) = 1$ holds on the nose
(\texttt{add\_compl\_eq\_one\_of\_complemented}).  On a connected
locale that fragment can be as small as $\{\bot, \top\}$.  In the
Boolean limit every element is complemented, the slack vanishes
identically, and Kolmogorov is recovered
(\texttt{classical\_additivity}, \texttt{classical\_slack\_zero}).

\paragraph{Mixtures.}
Valuations are closed under finite convex combination
(\texttt{Valuation.mix}), with every axiom inherited termwise,
because modularity is linear.  This innocuous fact becomes a
characterization in Section~\ref{sec:representation}.
