\section{Introduction}
\label{sec:intro}

Step-indexed reasoning has become the standard semantic technique
for verifying higher-order programming languages with mutable state,
recursive types, and concurrency. From Appel and McAllester's
original step-indexed model of recursive types
\cite{appel-mcallester2001} to the modern Iris
ecosystem~\cite{jung-et-al-2018}, the same basic idea recurs: instead
of asking whether a program satisfies a specification, ask whether it
satisfies the specification \emph{at every finite observation depth
$n$}, then recover the unindexed statement as the limit. Each level
of the hierarchy of approximations is defined without circularity by
appeal only to lower levels, and a ``later'' modality $\triangleright$
mediates the descent from one level to the next.

What the step-indexed literature does not usually say is that this
technique is, in classical set-theoretic terms, \emph{forcing}. Cohen
introduced forcing in 1963 to prove the independence of the continuum
hypothesis~\cite{cohen1963-independence-i, cohen1966}. In its
original presentation --- called \emph{ramified forcing} after the
ramified type theory of Russell and
Whitehead~\cite{russell-whitehead-PM} --- the construction of the
generic extension is stratified by a well-founded hierarchy of
\emph{levels}, with recursion at level $\alpha$ permitted to consult
only levels strictly below $\alpha$. Shoenfield's 1971
reformulation~\cite{shoenfield1971} eliminated the ramification for
the set-theoretic application: the modern textbook presentation of
forcing does not use it. But the structure --- well-founded
stratification, level descent, a forcing relation
$p \Vdash \varphi$ that is monotone in the strength of $p$ --- is
exactly the structure of step-indexed logic.

In this paper we make the correspondence precise. The forcing
\emph{conditions} of step-indexed Hoare logic are the natural numbers
$\omega$ with the order $\le$; the forcing \emph{relation}
$n \Vdash \varphi$ is monotone in $n$ (downward closure); the
\emph{later} modality $\triangleright$ is the level descent operator
of ramified forcing; and Löb's rule, which underwrites all
recursive reasoning in step-indexed systems, is precisely
well-founded induction on the forcing order, internalized into the
logic.

\paragraph{What is new here.}
Two things, both mechanized.  The first is a structural
\emph{diagnostic} for when the internal Löb rule is the right
soundness tool for a recursive Hoare-style rule.  The diagnostic is
sharp and binary, and it cuts on what we call the soundness
predicate's \emph{recursion locus}.  In the IMP setting
(Section~\ref{sec:imp}), the \texttt{while} rule's soundness predicate
is recursive only in the step index; an outer induction on the step
index produces the same proof as an appeal to the internal Löb rule,
and the two are mechanizably interchangeable.  We exhibit both side
by side (\texttt{hoare\_while} and \texttt{hoare\_while\_iLob}).  In
the $\lambda$-calculus-with-\texttt{fix} setting
(Section~\ref{sec:lam}), the \texttt{fix} rule's soundness predicate
is recursive in the program itself; an outer induction on the step
index fights this structure, and the natural proof invokes the
internal Löb rule directly, with no meta-level induction on $n$
anywhere in the proof.  Section~\ref{sec:imp-vs-lam} makes the
diagnostic precise.  This articulation does not, to our knowledge,
appear in either the synthetic guarded domain theory literature
(where the topos-of-trees account makes the Löb rule available but
does not distinguish the IMP-style and $\lambda$-fix-style settings)
or the Iris literature (where internal Löb is heavily used but the
question of when it is essential versus optional is not explicitly
diagnosed).

The second new piece is a \emph{negative result}: the
well-foundedness of the forcing notion is essential.
Section~\ref{sec:lob-essential} exhibits a would-be forcing structure
over the integers $\mathbb{Z}$, in which the carrier and the order
satisfy every part of the abstract framework except well-foundedness.
All of \texttt{iProp}, the Heyting connectives, the implication, and
the later modality go through.  But Löb's rule does not: the premise
$(\triangleright \bot) \to \bot$ is forced at every condition ---
vacuously, since $\mathbb{Z}$ has no minimum --- while $\bot$ is
forced at none, so the entailment fails.  This is a step-indexed
Hoare-logic instance of a classical fact about Gödel--Löb modal
logic~\cite{smorynski1985,boolos1993}.  In that classical form it is
folklore.  Brought into the verification setting and mechanized, as
far as we can determine, it has not previously appeared, and it
states a soundness desideratum that any forcing-style step-indexed
semantics has to meet: \emph{the well-founded forcing notion is
constitutive of the modal logic, not a presentation choice that can
be lifted away by a clever recursion on rank}.

\paragraph{The framework that supports these observations.}
Both observations rest on a re-presentation of step-indexed Hoare
logic as forcing, mechanized in Rocq across a propositional layer
(Section~\ref{sec:prop}), Hoare logic for IMP (Section~\ref{sec:imp})
and for an untyped $\lambda$-calculus with \texttt{fix}
(Section~\ref{sec:lam}), and an abstract framework parametric in an
arbitrary well-founded forcing notion (Section~\ref{sec:abstract}).
The abstract framework instantiates at $(\omega, \le)$ to recover the
standard step-indexed semantics and at $(\omega \times \omega,
\le_{\mathrm{pw}})$ to give a non-trivially different second example.
In the abstract framework, Löb's rule is proved by
\texttt{well\_founded\_ind} on the strict refinement relation ---
making precise the equivalence between the internal Löb rule and
well-founded induction on the forcing order.

The re-presentation is not itself novel.  The topos-of-trees account
of step-indexing~\cite{birkedal-et-al-2012} is implicitly a
forcing-style semantics in the model-theoretic sense
(Section~\ref{sec:topos}); the Iris ecosystem~\cite{jung-et-al-2018}
uses internal Löb as a workhorse without referring to it as a
forcing construction; and the Jaber--Tabareau--Sozeau forcing
translation of CIC~\cite{jaber-tabareau-sozeau-2012} produces, at
the level of propositions, structures that coincide with our
\texttt{iProp} (Section~\ref{sec:bridge}).  The framework we give is
the smallest such account that supports the two technical
observations stated above; its purpose is not to compete with these
existing constructions, but to expose the structure inside them that
makes the diagnostic and the counterexample stateable in the first
place.

\paragraph{Contributions.}

\begin{enumerate}

\item \textbf{The recursion-locus diagnostic}
(Section~\ref{sec:imp-vs-lam}): an explicit, mechanized
distinction between IMP-style first-order recursive Hoare rules
(where internal Löb and meta-level induction on the step index are
interchangeable) and $\lambda$-fix-style higher-order recursive
Hoare rules (where the soundness predicate is itself recursive in
the program and only internal Löb plays to that structure).  We
mechanize both forms of \texttt{hoare\_while} for IMP and the
\texttt{wp\_fix} rule for the $\lambda$-calculus, and exhibit the
recursion-locus distinction between them.

\item \textbf{A non-well-founded counterexample to Löb's rule}
(Section~\ref{sec:lob-essential}): a mechanized would-be forcing
structure over $\mathbb{Z}$ in which every iProp construction goes
through and the premise $(\triangleright \bot) \to \bot$ is forced
everywhere, but $\bot$ is forced nowhere, so Löb's rule fails.

\item \textbf{A mechanized forcing-style re-presentation of
step-indexed Hoare logic} (Sections~\ref{sec:prop}--\ref{sec:lam}):
the substrate that supports the two observations above.  Covers a
forcing-style propositional semantics with the later modality and
Löb's rule, Hoare logic over IMP, and Hoare logic over an untyped
$\lambda$-calculus with \texttt{fix}.

\item \textbf{An abstract forcing framework
parametric in the forcing notion} (Section~\ref{sec:abstract}),
instantiated at $(\omega, \le)$ to recover the standard
step-indexed semantics and at $(\omega \times \omega, \le_{\mathrm{pw}})$
to demonstrate genericity.  In the abstract setting Löb's rule is
proved by \texttt{well\_founded\_ind}.

\item \textbf{A formalized embedding of meta-level $\mathsf{Prop}$
into the propositional fragment of the Jaber--Tabareau--Sozeau
forcing translation of CIC}
(Section~\ref{sec:bridge})~\cite{jaber-tabareau-sozeau-2012}.  The
embedding is a Heyting algebra homomorphism; the later modality
$\triangleright$ is the step-indexed-specific content lying outside
its image.  A full mechanization of the type-level translation is
future work.

\end{enumerate}

\paragraph{Scope.}
We work at the level of plain Hoare triples and a partial-correctness
wp predicate.  We do not develop a separation logic, do not model a
heap, and do not address concurrency.  The $\lambda$-calculus of
Section~\ref{sec:lam} is untyped and the recursive spec we use takes
a single predicate $S$ as both pre- and post-condition; the rule
generalizes to distinct pre/post in the formalization but the
demonstrative examples use the single-predicate form.  Our
contribution is therefore conceptual --- it makes precise a
soundness-machinery claim about step-indexed Hoare logic --- not a
verification-engineering claim about reaching Iris-scale
applicability.  Whether the forcing reading scales gracefully to
Iris-style separation logic, resource state, and concurrent step
counters is the question the abstract framework of
Section~\ref{sec:abstract} is intended to make tractable; we do not
answer it here.

\paragraph{Outline.}
Section~\ref{sec:prop} gives the propositional layer of the forcing
semantics, including the Heyting-algebra connectives, the later
modality, and Löb's rule. Section~\ref{sec:imp} develops Hoare logic
for IMP and proves the \texttt{while} rule by meta-level induction;
Section~\ref{sec:lam} develops Hoare logic for an untyped
$\lambda$-calculus with \texttt{fix} and proves the soundness rule
for \texttt{fix} via the internal Löb rule. Section~\ref{sec:imp-vs-lam}
compares these two proofs explicitly and articulates the diagnostic
claim. Section~\ref{sec:abstract} gives the abstract forcing
framework. Section~\ref{sec:lob-essential} proves Löb's rule fails
without well-foundedness. Section~\ref{sec:bridge} formalizes the
embedding of meta-level $\mathsf{Prop}$ into the propositional
fragment of the JTS forcing translation.
Section~\ref{sec:related} positions the contribution against the
literature. Section~\ref{sec:conclusion} concludes.

\paragraph{Rocq development.}
The complete Rocq development is approximately 800 lines across 11
files, building cleanly against the Rocq standard library only (no
external libraries such as Iris or stdpp).

