%% =====================================================================
%% main.tex  --- Step-Indexed Hoare Logic as Ramified Forcing
%% Draft for CPP submission (acmart sigplan).
%%
%% To compile:
%%   pdflatex main && bibtex main && pdflatex main && pdflatex main
%% Or: make
%%
%% This file inlines the related-work and introduction/conclusion drafts
%% (see related-work.tex and intro-concl.tex for the standalone versions).
%% Body sections marked [TBD] are still to be written.
%% =====================================================================

\documentclass[sigplan,review,anonymous,screen]{acmart}

\settopmatter{printacmref=false}
\setcopyright{none}
\renewcommand\footnotetextcopyrightpermission[1]{}
\pagestyle{plain}

\usepackage{amsmath,amssymb,amsthm}
\usepackage{listings}
\usepackage{microtype}

\acmConference[CPP '27]{Certified Programs and Proofs}{January 2027}{(submitted)}
\acmYear{2027}
\acmISBN{}
\acmDOI{}

\citestyle{acmauthoryear}

\theoremstyle{plain}
\newtheorem{theorem}{Theorem}
\newtheorem{lemma}[theorem]{Lemma}
\newtheorem{proposition}[theorem]{Proposition}
\newtheorem{corollary}[theorem]{Corollary}

\theoremstyle{definition}
\newtheorem{definition}[theorem]{Definition}
\newtheorem{example}[theorem]{Example}

\theoremstyle{remark}
\newtheorem{remark}[theorem]{Remark}

\begin{document}

\title{Step-Indexed Hoare Logic as Ramified Forcing}

\author{Anonymous}
\affiliation{%
  \institution{Anonymous}
  \country{}}
\email{anonymous@example.com}

\begin{abstract}
We show that step-indexed Hoare logic is, conceptually and technically,
\emph{ramified forcing} --- Cohen's original forcing construction
stratified by a well-founded hierarchy of levels. The
``later'' modality $\triangleright$ is ramification's level-descent
operator; Löb's rule, the central technical device of step-indexed
recursive reasoning, is precisely well-founded induction on the
forcing order, internalized into the logic. We make the correspondence
precise via a Rocq mechanization that develops the forcing-style
propositional semantics, builds Hoare logic over IMP and over an
untyped $\lambda$-calculus with general recursion, gives an abstract
framework parametric in the forcing notion, proves that the
well-foundedness is essential by exhibiting a non-well-founded
counterexample over $\mathbb{Z}$ where Löb's rule fails, and
formalizes the bridge to the Jaber--Tabareau--Sozeau forcing
translation of CIC via a Heyting-algebra-homomorphism embedding
$\mathsf{Prop} \hookrightarrow \mathtt{iProp}$. A central methodological
observation is that for higher-order recursion the natural soundness
proof goes through the internal Löb rule directly, whereas for
first-order IMP the same content can be obtained either via the
internal rule or via meta-level induction on the step index. We
mechanize both forms side by side to make the equivalence concrete.
The complete development is approximately 800 lines of Rocq across
11 files, with no external dependencies.
\end{abstract}

\begin{CCSXML}
<ccs2012>
   <concept>
       <concept_id>10003752.10003790.10011740</concept_id>
       <concept_desc>Theory of computation~Modal and temporal logics</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
   <concept>
       <concept_id>10003752.10010124.10010138</concept_id>
       <concept_desc>Theory of computation~Program semantics</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
   <concept>
       <concept_id>10003752.10003790.10011742</concept_id>
       <concept_desc>Theory of computation~Type theory</concept_desc>
       <concept_significance>300</concept_significance>
   </concept>
   <concept>
       <concept_id>10003752.10010124.10010138.10010143</concept_id>
       <concept_desc>Theory of computation~Hoare logic</concept_desc>
       <concept_significance>500</concept_significance>
   </concept>
</ccs2012>
\end{CCSXML}

\ccsdesc[500]{Theory of computation~Modal and temporal logics}
\ccsdesc[500]{Theory of computation~Program semantics}
\ccsdesc[500]{Theory of computation~Hoare logic}
\ccsdesc[300]{Theory of computation~Type theory}

\keywords{step-indexed semantics, forcing, ramified forcing, Hoare logic,
  Löb's rule, Rocq mechanization, Gödel--Löb logic, forcing translations}

\maketitle

%% =====================================================================
\section{Introduction}
\label{sec:intro}

Step-indexed reasoning has become the standard semantic technique
for verifying higher-order programming languages with mutable state,
recursive types, and concurrency. From Appel and McAllester's
original step-indexed model of recursive types
\cite{appel-mcallester2001} to the modern Iris
ecosystem~\cite{jung-et-al-2018}, the same basic idea recurs: instead
of asking whether a program satisfies a specification, ask whether it
satisfies the specification \emph{at every finite observation depth
$n$}, then recover the unindexed statement as the limit. Each level
of the hierarchy of approximations is defined without circularity by
appeal only to lower levels, and a ``later'' modality $\triangleright$
mediates the descent from one level to the next.

What the step-indexed literature does not usually say is that this
technique is, in classical set-theoretic terms, \emph{forcing}. Cohen
introduced forcing in 1963 to prove the independence of the continuum
hypothesis~\cite{cohen1963-independence-i, cohen1966}. In its
original presentation --- called \emph{ramified forcing} after the
ramified type theory of Russell and
Whitehead~\cite{russell-whitehead-PM} --- the construction of the
generic extension is stratified by a well-founded hierarchy of
\emph{levels}, with recursion at level $\alpha$ permitted to consult
only levels strictly below $\alpha$. Shoenfield's 1971
reformulation~\cite{shoenfield1971} eliminated the ramification for
the set-theoretic application: the modern textbook presentation of
forcing does not use it. But the structure --- well-founded
stratification, level descent, a forcing relation
$p \Vdash \varphi$ that is monotone in the strength of $p$ --- is
exactly the structure of step-indexed logic.

In this paper we make the correspondence precise. The forcing
\emph{conditions} of step-indexed Hoare logic are the natural numbers
$\omega$ with the order $\le$; the forcing \emph{relation}
$n \Vdash \varphi$ is monotone in $n$ (downward closure); the
\emph{later} modality $\triangleright$ is the level descent operator
of ramified forcing; and Löb's rule, which underwrites all
recursive reasoning in step-indexed systems, is precisely
well-founded induction on the forcing order, internalized into the
logic.

\paragraph{Why does this matter?}
The forcing reading is not just a re-labelling. It clarifies three
things that are otherwise scattered across the step-indexed
literature.

First, it explains \emph{why} the technique has the shape it does.
The well-founded stratification is not a presentation convenience to
be eliminated; it is constitutive of the soundness of the modal
logic. Section~\ref{sec:lob-essential} makes this precise: we exhibit
a non-well-founded forcing notion (the integers $\mathbb{Z}$) where
every other piece of the framework goes through, but Löb's rule
fails.

Second, it diagnoses where the internal Löb rule earns its keep.
Section~\ref{sec:imp-vs-lam} compares two mechanized soundness proofs
of recursive Hoare rules: the \texttt{while} rule for IMP
(Section~\ref{sec:imp}) and the \texttt{fix} rule for the untyped
$\lambda$-calculus (Section~\ref{sec:lam}). The first admits an
unproblematic meta-level proof by induction on the step index; the
second does not, and the natural proof invokes the internal Löb rule
directly. The structural reason --- that the soundness predicate is
itself recursive in the \texttt{fix} case --- is made precise by the
forcing reading.

Third, it identifies step-indexed Hoare logic as the propositional
fragment of an existing construction in the type-theory literature:
the Jaber--Tabareau--Sozeau forcing translation of
CIC~\cite{jaber-tabareau-sozeau-2012}. Section~\ref{sec:bridge}
formalizes this identification via a constant embedding
$\mathsf{Prop} \hookrightarrow \mathtt{iProp}$ that is a Heyting
algebra homomorphism, with $\triangleright$ appearing as the
step-indexed-specific addition.

\paragraph{Contributions.}

\begin{enumerate}

\item A mechanized re-presentation of step-indexed Hoare logic as
ramified forcing, in Rocq. The propositional layer
(Section~\ref{sec:prop}) gives a forcing semantics for an
intuitionistic logic with the later modality and Löb's rule. The
Hoare layer is constructed over IMP (Section~\ref{sec:imp}) and over
an untyped $\lambda$-calculus with general recursion
(Section~\ref{sec:lam}).

\item An explicit comparison of the IMP \texttt{while} rule and the
$\lambda$-calculus \texttt{fix} rule that diagnoses where the
internal Löb rule is essential (Section~\ref{sec:imp-vs-lam}). For
IMP, soundness goes through either via meta-level induction on the
step index or via the internal Löb rule; we exhibit both proofs and
observe their structural identity. For the $\lambda$-calculus with
\texttt{fix}, the natural soundness proof must invoke the internal
Löb rule, because the soundness predicate is itself recursive in the
program. This is the strongest piece of evidence for the slogan that
``Löb is what makes step-indexed Hoare logic work in higher-order
settings.''

\item An abstract forcing framework (Section~\ref{sec:abstract})
parametric over an arbitrary well-founded forcing structure. We
exhibit two instances: $(\omega, \le)$, recovering the standard
step-indexed semantics; and $(\omega \times \omega, \le_{\mathrm{pw}})$,
with the pointwise order, modelling independent step-indices for
parallel components. The internal Löb rule, in the abstract setting,
is proved by \texttt{well\_founded\_ind} on the strict refinement
order. This makes precise the equivalence: \emph{internal Löb is
well-founded induction on the forcing order, internalized}.

\item A small new metatheoretic result: \emph{well-foundedness of the
forcing notion is essential} (Section~\ref{sec:lob-essential}). We
exhibit a non-well-founded ``would-be'' forcing structure over
$\mathbb{Z}$, in which every iProp construction goes through except
Löb's rule. The premise of Löb's rule $(\triangleright \bot) \to \bot$
is forced at every condition --- vacuously, since $\mathbb{Z}$ has no
minimum --- while $\bot$ is forced at none. This is a special case of
the model theory of the Gödel--Löb provability
logic~\cite{solovay1976,boolos1993}, brought into the step-indexed
Hoare logic setting and mechanized.

\item A partial bridge to the Jaber--Tabareau--Sozeau forcing
translation of CIC~\cite{jaber-tabareau-sozeau-2012}
(Section~\ref{sec:bridge}). We formalize the constant embedding
$\mathsf{Prop} \hookrightarrow \mathtt{iProp}$ as a Heyting algebra
homomorphism, and observe that the $\triangleright$ modality is the
genuinely new content of the iProp universe over the embedded image
of Prop. A full mechanization of the type-level JTS translation is
future work.

\end{enumerate}

\paragraph{What is and is not new.}
The individual ingredients are largely known to specialists in
neighbouring fields. The topos-of-trees perspective on
step-indexing~\cite{birkedal-et-al-2012} is implicitly a forcing
semantics; the Gödel--Löb modal logic of provability is the modal
logic of provability and well-foundedness; the JTS forcing
translation of CIC is the type-theoretic analog of forcing. What is
new is the assembly: a single mechanized account that makes the
forcing reading explicit, diagnoses the role of the internal Löb
rule, and gives a parametric framework with the metatheoretic
guarantee that well-foundedness cannot be relaxed.

\paragraph{Outline.}
Section~\ref{sec:prop} gives the propositional layer of the forcing
semantics, including the Heyting-algebra connectives, the later
modality, and Löb's rule. Section~\ref{sec:imp} develops Hoare logic
for IMP and proves the \texttt{while} rule by meta-level induction;
Section~\ref{sec:lam} develops Hoare logic for an untyped
$\lambda$-calculus with \texttt{fix} and proves the soundness rule
for \texttt{fix} via the internal Löb rule. Section~\ref{sec:imp-vs-lam}
compares these two proofs explicitly and articulates the diagnostic
claim. Section~\ref{sec:abstract} gives the abstract forcing
framework. Section~\ref{sec:lob-essential} proves Löb's rule fails
without well-foundedness. Section~\ref{sec:bridge} establishes the
constant-embedding bridge to JTS forcing translations.
Section~\ref{sec:related} positions the contribution against the
literature. Section~\ref{sec:conclusion} concludes.

\paragraph{Rocq development.}
The complete Rocq development is approximately 800 lines across 11
files, building cleanly against the Rocq standard library only (no
external libraries such as Iris or stdpp).

%% =====================================================================
\section{The Propositional Layer}
\label{sec:prop}

\noindent\textit{[TBD: forcing semantics for propositional
intuitionistic logic with the later modality and Löb's rule. Covers:
conditions as $\mathbb{N}$ with $\le$; \texttt{iProp} as
downward-closed predicates with monotonicity; Heyting connectives
$\top, \bot, \wedge, \vee, \to$; the later modality $\triangleright$;
Löb's rule $(\triangleright\varphi \to \varphi) \vdash \varphi$, proved
by induction on the step index; the sanity check that
$\triangleright\bot \not\vdash \bot$. References Phase 1 of the Rocq
development.]}

%% =====================================================================
\section{Hoare Logic for IMP}
\label{sec:imp}

\noindent\textit{[TBD: the first-order calibration setting. Covers the
IMP language and its small-step semantics; the step-indexed
\texttt{wp\_at} predicate and the Hoare triple as an
\texttt{iProp}; soundness of \texttt{skip}, assignment, sequencing,
conditional, and consequence. Then the \texttt{while} rule, proved
by induction on the step index, and an alternative proof invoking
the internal Löb rule via \texttt{ientails\_trans} and \texttt{iLob}.
References Phases 2a/2b/2c.]}

%% =====================================================================
\section{Hoare Logic for an Untyped $\lambda$-Calculus with Recursion}
\label{sec:lam}

\noindent\textit{[TBD: the higher-order setting where step-indexing
becomes essential. Covers the untyped $\lambda$-calculus with
\texttt{fix} (de Bruijn syntax, CBV small-step semantics with
$\beta$ and fix-unfolding); the step-indexed \texttt{wp\_at} with
an explicit safety conjunct (because the $\lambda$-calculus has
stuck terms); the \texttt{wp\_value} and \texttt{wp\_pure\_step}
rules. The \texttt{wp\_fix} soundness rule, proved by
\texttt{apply ientails\_trans} to reduce to the form
$\top \vdash \triangleright\varphi \to \varphi$, then \texttt{apply
iLob}. No meta-level induction. The worked example
\texttt{recursive\_id} (a fix-based identity) and \texttt{bottom}
(canonical divergent fix-point). References Phases 3a/3b/3c/3d.]}

%% =====================================================================
\section{Why the Two Proofs Look Different}
\label{sec:imp-vs-lam}

\noindent\textit{[TBD: the central diagnostic section. Compares the
two soundness proofs of Sections~\ref{sec:imp} and~\ref{sec:lam} side
by side. In IMP, the wp predicate for \texttt{while} reduces to an
induction on the step count without further recursion in the
predicate itself. In the $\lambda$-calculus with \texttt{fix}, the
wp predicate for \texttt{fix} is intrinsically recursive in the
self-reference, and unrolling the recursion at the meta level
requires the (well-founded) induction principle that Löb's rule
internalizes. The structural diagnostic: \emph{when the soundness
predicate is itself recursive in the program, internal Löb is the
natural tool; otherwise it is interchangeable with outer induction
on the step index.}]}

%% =====================================================================
\section{An Abstract Forcing Framework}
\label{sec:abstract}

\noindent\textit{[TBD: Phase 4. The
\texttt{ForcingStructure} record packaging a preorder $\le$, a strict
relation $<$, the standard order axioms, and well-foundedness of
$<$. Over an arbitrary \texttt{ForcingStructure}, the
\texttt{iProp}, Heyting connectives, and the later modality
$\triangleright\varphi$ defined as universal-over-strict-predecessors:
$p \Vdash \triangleright\varphi$ iff $\forall p' \prec p, p' \Vdash
\varphi$. Löb's rule proved by \texttt{well\_founded\_ind} on
$<$. Two instances: \texttt{nat\_FS} (the standard $(\omega, \le)$
recovering Phase 1) and \texttt{prod\_FS} (the pointwise product
$\omega \times \omega$). The conceptual claim: at the meta level,
internal Löb's rule \emph{is} well-founded induction on the forcing
order, viewed internally.]}

%% =====================================================================
\section{Well-foundedness is Essential}
\label{sec:lob-essential}

\noindent\textit{[TBD: Phase 5. The metatheoretic observation that
the $\texttt{fs\_lt\_wf}$ axiom in \texttt{ForcingStructure} cannot
be dropped. We construct a ``would-be'' iProp framework over
$\mathbb{Z}$, in which all of \texttt{iPropZ},
\texttt{iImplZ}, \texttt{iLaterZ} go through with their standard
definitions. We show: (i) the premise of Löb's rule,
$(\triangleright \bot) \to \bot$, is forced at every condition,
because $\mathbb{Z}$ has no minimum and so $\triangleright \bot$ is
forced at no condition; (ii) $\bot$ itself is not forced at every
condition; (iii) hence the entailment
$(\triangleright \bot \to \bot) \vdash \bot$ does not hold.
This is the step-indexed analog of the modal logic fact that GL is
sound on the class of converse-well-founded Kripke frames and not
on the integers. The methodological upshot: the ``ramification'' in
ramified forcing is not a presentation convenience; it is
constitutive of the modal logic.]}

%% =====================================================================
\section{Bridge to Forcing Translations}
\label{sec:bridge}

\noindent\textit{[TBD: Phase 6. The constant embedding $\mathsf{Prop}
\hookrightarrow \mathtt{iProp}$ defined as
$\lceil P \rceil := \texttt{MkProp}\, (\lambda \_, P)\, \cdots$. We
prove it is a Heyting algebra homomorphism: it commutes with $\top$,
$\bot$, $\wedge$, $\vee$, $\to$ in both directions of entailment.
The genuine new content of \texttt{iProp} over the embedded image is
the $\triangleright$ modality, witnessed by
$\triangleright \lceil \mathrm{False} \rceil \not\vdash
\lceil \mathrm{False} \rceil$. The framework of
Section~\ref{sec:abstract} is therefore the propositional fragment of
the Jaber--Tabareau--Sozeau forcing translation of CIC
\cite{jaber-tabareau-sozeau-2012}, instantiated at an arbitrary
forcing structure. A full mechanization of the type-level JTS
translation is future work.]}

%% =====================================================================
\section{Related Work}
\label{sec:related}

This section is organized into seven literature threads:
(i) step-indexing in programming language semantics; (ii) forcing in
set theory, ramified and unramified; (iii) topos-theoretic models of
step-indexing; (iv) forcing translations in type theory; (v) the
modal logic of provability (\textsf{GL}) and well-foundedness;
(vi) constructive modal logic for guarded recursion; (vii)
predicativity and ramified hierarchies in foundations of
mathematics. A final subsection states the gap this paper fills.

\subsection{Step-indexing in programming language semantics}

Step-indexed logical relations were introduced by Appel and
McAllester \cite{appel-mcallester2001} to give a semantic account of
recursive types in the setting of foundational proof-carrying code.
The technique gives a meaning to recursive types not via a fixed
point construction in some semantic domain, but by approximating the
intended meaning at every finite ``observation depth'' $n$. Two terms
are equivalent at index $n$ if they cannot be distinguished by any
context running for at most $n$ steps; the desired equivalence is then
the limit (intersection) of all $n$-equivalences. Appel and McAllester
show that this gives a sound model for recursive types and that the
resulting logical relation is preserved under reduction --- the
canonical \emph{step-down} property that we will recognise as the
monotonicity of step-indexed propositions.

Ahmed's PhD thesis \cite{ahmed2004} extended step-indexing to
languages with mutable references, including general references that
can store functions of any type (``higher-order store''). This is the
classical setting in which step-indexing is essential rather than
merely convenient: the standard set-theoretic recursion on type
structure is blocked by the circularity that comes from storing
higher-order functions in mutable cells. Ahmed shows how the
step-indexed approach absorbs this circularity into the index
hierarchy.

Birkedal and collaborators have extended these ideas in two
directions. In one direction, \emph{topos-theoretic} reformulations
(see Section~\ref{sec:topos} below) replaced the explicit step
indices with categorical structure. In the other, the technique was
scaled up to richer programming languages and verification settings;
see Birkedal et al.\
\cite{birkedal-et-al-2011-stepindexed} on step-indexed Kripke models
for capability calculi.

The current state of the art on the programming-languages side is
the \emph{Iris} family of separation logics
\cite{jung-et-al-2018,krebbers-jung-bizjak-birkedal-dreyer-2017}.
Iris extends step-indexed separation logic with higher-order ghost
state, invariants, and a powerful set of derived rules. Its
soundness theorem is mechanized in Coq against a model built
ultimately on step-indexed propositions. The internal logic of Iris
is itself a step-indexed modal logic with a ``later'' modality
$\triangleright$ and a Löb rule; substantial libraries of Coq
tactics support reasoning within it. Other recent step-indexed
systems include VST \cite{appel2014vst} and Transfinite Iris
\cite{spies2021transfinite}.

Two technical observations from this literature are essential for
our work. First, the ``later'' modality has been a unifying device
since at least Appel et al.\ \cite{appel-melham-mcallester2007}, who
introduced it as the very-modal $\triangleright$ for guarded
recursion. Second, the soundness proof of every step-indexed logic
relies on a Löb-style argument --- either as an explicit lemma in
the internal logic, or implicitly through induction on the step index
in the soundness proof. Our paper takes seriously the question of
which formulation is more natural in which setting, and argues that
for higher-order recursion the internal Löb form is strictly
preferable.

\subsection{Forcing in set theory: ramified and unramified}

Cohen introduced forcing in his 1963 papers
\cite{cohen1963-independence-i, cohen1963-independence-ii} to prove
the independence of the continuum hypothesis from \textsf{ZFC}, and
gave a book-length treatment in 1966 \cite{cohen1966}. The original
presentation is what we will call \emph{ramified forcing}: the
construction of the generic extension $M[G]$ is stratified by ordinal
levels mimicking Gödel's constructible hierarchy. A ``ramified
language'' $\mathcal{L}^*$ has a type symbol for each ordinal
$\alpha$ below some sufficient bound, and a variable of type $\alpha$
ranges only over sets at levels strictly below $\alpha$. The forcing
relation $p \Vdash_\alpha \varphi$ is defined level by level: forcing
of a level-$\alpha$ formula recurses only on forcing at levels
strictly below $\alpha$.

The two features that make this presentation ``ramified'' are
predicative stratification (a level-$\alpha$ formula only quantifies
over levels strictly below $\alpha$) and explicit level descent (the
recursive definition of forcing is structured by ordinal predecessor).
Both features are inherited from the ramified theory of types of
Russell and Whitehead \cite{russell-whitehead-PM} and from the broader
tradition of predicative analysis \cite{weyl1918,feferman1998}.

Shoenfield's 1971 lecture \cite{shoenfield1971} reformulated forcing
without the ramification: forcing conditions and $P$-names are
defined by transfinite recursion on rank, and the forcing relation is
defined by recursion on formula complexity, without explicit
stratification by syntactic level. Shoenfield's presentation is
mathematically equivalent to Cohen's --- the same generic extensions
are produced --- but it is simpler, and it has become the standard
modern presentation in textbooks \cite{kunen2011,jech2003}.
Boolean-valued models, due to Scott and Solovay
\cite{scott-solovay1967}, give yet another equivalent presentation.

What is preserved across these reformulations is the underlying
\emph{forcing relation} $p \Vdash \varphi$ together with its
fundamental properties. What was eliminated --- the explicit ramified
syntactic hierarchy --- is exactly what we will argue is \emph{not}
eliminable in the step-indexed setting.

\subsection{Topos-theoretic models of step-indexing}
\label{sec:topos}

The connection between step-indexing and category-theoretic
constructions was made explicit by Birkedal, Møgelberg,
Schwinghammer, and Støvring \cite{birkedal-et-al-2012} via the
\emph{topos of trees}, the presheaf topos on the partially ordered
set $(\omega, \le)$. They observed that step-indexed propositions are
the propositions of the internal logic of this topos, and that the
``later'' modality $\triangleright$ arises naturally as a particular
endofunctor on objects of the topos. Their paper introduces
\emph{synthetic guarded domain theory}, in which all of the
step-indexed semantics is internalized: rather than indexing
manually, one works in a type theory whose intrinsic semantics is
the topos of trees.

The presheaf-topos perspective makes precise much of what the early
step-indexed literature handled informally. In particular, the
monotonicity condition on step-indexed propositions corresponds
exactly to the presheaf condition (functoriality), and the soundness
of the Löb rule is the existence of a fixed point of the
$\triangleright$ functor (the unique fixed point of $\triangleright X
= F(X)$ for a contractive $F$, or equivalently the well-foundedness
of $(\omega, <)$ as the index domain). Subsequent work has refined
the categorical setting and used it to model more sophisticated
languages \cite{bizjak-grathwohl-clouston-mogelberg-birkedal-2016}.

The presheaf-topos perspective is implicitly a forcing-style
semantics in the model-theoretic sense: a presheaf topos on
$(P, \le)$ is exactly the topos of sheaves with respect to the
trivial Grothendieck topology on $P$, and these are exactly the
Kripke models of intuitionistic logic with worlds in $P$. The
topos's truth value object plays the role of a Heyting algebra-valued
``forcing'' truth value. To our knowledge, this connection has not
been emphasized in the topos-of-trees literature, and the word
``forcing'' has not been prominent in it. Our paper makes this
terminological and conceptual connection explicit.

\subsection{Forcing translations in type theory}
\label{sec:jts}

A parallel literature, originating in the constructive type theory
community, develops syntactic \emph{forcing translations} of type
theories. The seminal work here is Jaber, Tabareau, and Sozeau
\cite{jaber-tabareau-sozeau-2012}, who give a forcing translation of
the Calculus of Constructions: given a forcing notion (a poset
$P$), they exhibit a syntactic translation that takes a CIC term
$M : T$ to a forced term $M^P : T^P$, where the type translation
$T \mapsto T^P$ corresponds to ``the forced version of $T$.'' For
propositions, $T^P$ is a downward-closed predicate on $P$ ---
exactly our \texttt{iProp} of Section~\ref{sec:prop}.

This line was extended in \cite{jaber-et-al-2016} (the
``definitional side of the forcing''), which gives a refined
translation that preserves more of the computational structure of
the original type theory. Coquand and collaborators have related
work on constructive presentations of forcing in type-theoretic
settings \cite{coquand-jaber2010}.

The Jaber et al.\ translations are the type-theoretic version of
ramified forcing for type theory: a parametric construction of a
``forced universe'' that supports the same operations as the
unforced one. Our Phase 4 abstract framework recovers the
propositional fragment of this translation by direct construction in
Rocq. The Phase 6 file makes the embedding precise via a constant
embedding $\mathsf{Prop} \hookrightarrow \mathtt{iProp}$ that is a
Heyting algebra homomorphism.

What our work does not do --- and what the Jaber et al.\ line does
--- is give a translation of the \emph{full} type theory, including
dependent types. A natural future direction is to formalize the
restriction of the Jaber et al.\ translation to the propositional
fragment in Rocq and check (mechanically) that it produces our
\texttt{iProp} up to definitional equality.

\subsection{Modal logic of provability and well-foundedness}
\label{sec:gl}

The Gödel--Löb provability logic \textsf{GL} is the modal logic
characterized by the axiom $\Box(\Box \varphi \to \varphi) \to
\Box \varphi$ (the ``Löb axiom''), together with the standard
$\mathsf{K}$ axioms. Solovay's arithmetical completeness theorem
\cite{solovay1976} establishes that \textsf{GL} is sound and complete
for arithmetical interpretations of $\Box$ as ``it is provable in
Peano Arithmetic that\ldots''. From the model-theoretic side,
\textsf{GL} is sound and complete for the class of transitive
Kripke frames whose accessibility relation $R$ is
\emph{converse well-founded} (no infinite ascending chain
$w_0 R w_1 R w_2 R \cdots$); see Smoryński \cite{smorynski1985} and
Boolos \cite{boolos1993} for textbook treatments.

The model theory of \textsf{GL} is directly relevant to our work.
The ``later'' modality $\triangleright$ of step-indexed logic is
formally exactly a \textsf{GL}-style provability $\Box$ once we
identify the strict refinement order $\prec$ on conditions with the
converse of the \textsf{GL} accessibility relation. Löb's rule of
step-indexed logic is the Löb axiom of \textsf{GL}. The fact that
$\triangleright$ requires well-foundedness for Löb's rule to hold
is the dual of \textsf{GL}'s requirement that the accessibility
relation be converse well-founded. Our Phase 5 result --- that Löb
fails over $\mathbb{Z}$ --- is therefore a special case of (the
converse of) the soundness of \textsf{GL} for converse-well-founded
frames.

What our paper adds to this picture is the observation that the
familiar PL-style step-indexed logic, including its mechanizations
in Iris and elsewhere, is in fact a propositional fragment of
\textsf{GL} indexed over a particular kind of \textsf{GL}-frame
(one with well-founded strict refinement). This connection is
folklore in modal logic but, as far as we are aware, has not been
made explicit or mechanized for the step-indexed Hoare logic
setting.

\subsection{Constructive modal logic and guarded recursion}

A separate line of work, with substantial overlap with synthetic
guarded domain theory, develops constructive modal logics for
guarded recursion. Atkey and McBride \cite{atkey-mcbride2013} use
the $\triangleright$ modality to characterize \emph{productive
coprogramming}: a corecursive function is productive iff each
recursive call sits under a $\triangleright$. Clouston, Bizjak,
Grathwohl, and Birkedal
\cite{clouston-et-al-2015,clouston-et-al-2017} develop guarded
$\lambda$-calculi with $\triangleright$ as a primitive type former,
supporting recursive types via a guarded fixed-point operator.
Møgelberg and others \cite{mogelberg2014,bizjak-mogelberg-2018}
extend these systems to dependent types and provide categorical
semantics.

This literature is closely allied with synthetic guarded domain
theory: both treat $\triangleright$ as a built-in modal operator and
exploit its compatibility with recursion. The work is structured
mostly around \emph{computational} and \emph{type-theoretic}
questions (``how do we type recursion productively?''), rather than
\emph{semantic} questions (``what does the step-indexed semantics
look like as a forcing construction?''). Our paper is complementary:
we work on the semantic side, with the forcing reading as our
organizing principle.

\subsection{Predicativity and ramified hierarchies}

The word ``ramified'' belongs to a tradition older than Cohen's
forcing. Russell and Whitehead \cite{russell-whitehead-PM}
introduced the ramified theory of types in \emph{Principia
Mathematica} to resolve the set-theoretic paradoxes via a
stratification of propositions by ``order,'' where a proposition of
order $n$ may quantify only over propositions of order $< n$. The
same idea animates Weyl's predicative analysis \cite{weyl1918}, in
which the real numbers are constructed within an arithmetical
hierarchy without appeal to impredicative set-theoretic principles.
Feferman's program of \emph{predicativity given the natural numbers}
\cite{feferman1998} develops this into a comprehensive foundation
for substantial parts of classical mathematics.

The thread connecting ramified type theory to ramified forcing is
that both stratify quantification by a well-founded hierarchy, and
both treat the hierarchy itself as part of the syntactic apparatus
rather than as a meta-level convenience. Shoenfield's unramification
of forcing showed that for forcing the stratification can be
absorbed into a single recursion on rank; what we argue, by
mechanizing step-indexed Hoare logic in higher-order settings, is
that for step-indexing the analogous unramification \emph{cannot} be
performed. The well-founded forcing notion is doing real work, not
just structuring the proof.

This connection also clarifies why the Iris-style internal logic
``feels predicative'': step-indexed propositions are stratified by a
well-founded hierarchy, and any recursive definition that does not
respect the hierarchy --- by mediating recursion through a
$\triangleright$ --- is rejected. The system is not just
predicative in spirit; it is precisely a ramified predicative system
in the technical sense.

\subsection{Hoare logic and program verification}

We omit a survey of classical Hoare logic and separation logic,
referring the reader to standard sources
\cite{hoare1969,ohearn-reynolds-yang-2001,brookes2007}. Two points
are worth noting. First, our IMP development in
Section~\ref{sec:imp} uses the plain (non-separation) version of
Hoare logic, since separation machinery is not needed for the
conceptual point we want to make about the \texttt{while} rule.
Second, Cook's relative completeness theorem \cite{cook1978} for
Hoare logic shows that the classical proof system is as complete as
can be hoped given any reasonable specification language; our
results are orthogonal to this --- we study how the soundness side
is best organized, not the proof system.

\subsection{The gap this paper fills}

The summaries above suggest that almost every ingredient of our work
is, individually, well-known to specialists in some adjacent
community. The topos-of-trees perspective on step-indexing is
familiar in synthetic guarded domain theory; the \textsf{GL} modal
logic of provability is familiar to modal logicians; the
ramified-forcing presentation of Cohen is familiar to set theorists;
the JTS forcing translation is familiar to the type theory
community; and the predicative reading of ramification is familiar
to foundations researchers.

What does \emph{not} appear in the literature, to our knowledge, is
a single mechanized account of step-indexed Hoare logic that
(i)~frames step-indexing explicitly as ramified forcing, making the
connection to Cohen's original presentation precise rather than
implicit in the topos-theoretic literature; (ii)~demonstrates that
internal Löb is the load-bearing soundness tool for higher-order
recursion by mechanizing both an IMP-style and a
$\lambda$-calculus-with-\texttt{fix} Hoare logic and exhibiting
both the meta-level and internal forms of the \texttt{while} rule's
soundness proof; (iii)~generalizes the framework to abstract
forcing notions parametrically, exhibiting at least one second
instance and recovering the standard $(\omega, \le)$ instance;
(iv)~proves the well-foundedness of the forcing notion is essential
by an explicit counterexample over $\mathbb{Z}$ where the iProp
constructions go through but Löb's rule fails; (v)~verifies the
bridge to the JTS forcing translation by formalizing the constant
embedding $\mathsf{Prop} \hookrightarrow \mathtt{iProp}$ as a
Heyting algebra homomorphism and identifying $\triangleright$ as the
step-indexed-specific addition.

The contribution is therefore not the introduction of any single
new construction but the \emph{integration} of several existing
threads into one mechanized account, plus the small new technical
result of Section~\ref{sec:lob-essential} and the explicit
comparison of Sections~\ref{sec:imp} and \ref{sec:lam} that
diagnoses where internal Löb earns its keep.

%% =====================================================================
\section{Conclusion}
\label{sec:conclusion}

We have given a forcing-style semantics for step-indexed Hoare logic,
mechanized in Rocq across the propositional layer, an IMP Hoare
logic, and an untyped $\lambda$-calculus with general recursion. The
forcing reading places the step-indexed technique in direct
correspondence with Cohen's ramified forcing construction: the
natural numbers are the forcing conditions, the later modality
$\triangleright$ is the level-descent operator, and Löb's rule is
well-founded induction on the forcing order, internalized into the
logic.

\paragraph{What we found.}
Two observations stand out. First, in the first-order setting of
IMP, the \texttt{while} rule admits two structurally identical
soundness proofs: an outer induction on the step index, and an
appeal to the internal Löb rule. The two proofs have the same
content because at the meta level, Löb's rule \emph{is} induction on
the step index. Mechanizing both side by side, as we do, makes the
equivalence concrete rather than analogical.

Second, in the higher-order setting of the untyped $\lambda$-calculus
with \texttt{fix}, the analogous soundness proof has a different
shape. The soundness predicate for a recursive function is itself
defined in terms of the function being verified, and unrolling this
self-reference at the meta level produces a non-trivial induction
that does not arise from the program's step count alone. The natural
proof, which we mechanize, invokes the internal Löb rule directly:
``show that under the assumption $\triangleright \varphi$ we have
$\varphi$, then conclude $\varphi$ by Löb.'' No meta-level induction
on a step count is needed. This is, in our opinion, the simplest
diagnostic for the slogan that \emph{Löb is what makes step-indexed
Hoare logic work for higher-order recursion}.

\paragraph{Three further observations.}
The abstract framework of Section~\ref{sec:abstract} makes precise
that the choice of $\omega$ as the step-index domain is a
\emph{computational} convenience rather than a foundational one. Any
well-founded forcing notion supports the same internal logic, with
$\triangleright$ defined uniformly as ``forces at every strict
predecessor'' and Löb derived by well-founded induction. We exhibit
the pointwise product $\omega \times \omega$ as a second instance, but
the framework is generic.

Section~\ref{sec:lob-essential} shows that the well-foundedness
cannot be dropped: we construct a non-well-founded forcing notion
over $\mathbb{Z}$ in which Löb's rule fails. This is a special case
of the model theory of \textsf{GL}, but the statement for step-indexed
Hoare logic does not, as far as we are aware, appear in the
verification literature. The claim has methodological weight:
step-indexed Hoare logic is not just ``Hoare logic with a step
counter''; it requires a well-founded forcing notion to be sound at
all.

Section~\ref{sec:bridge} verifies that the iProp framework is the
propositional fragment of the Jaber--Tabareau--Sozeau forcing
translation of CIC, via a constant embedding
$\mathsf{Prop} \hookrightarrow \mathtt{iProp}$ that we prove to be a
Heyting algebra homomorphism. The $\triangleright$ modality is the
genuinely new content of iProp over the embedded image, and arises
uniformly from the well-founded strict refinement on conditions.

\paragraph{Limitations.}
The Hoare logic for the $\lambda$-calculus in
Section~\ref{sec:lam} is, at present, restricted to partial
correctness against a single specification predicate $S$ that
serves as both pre- and post-condition. The fix soundness rule
\texttt{wp\_fix} (Section~\ref{sec:lam}) accordingly handles only
this case. A more general rule with distinct pre/post-conditions, a
rule for fix at higher types, and an integration with separation
logic on a heap are all natural directions but lie outside the
present development.

The bridge to JTS forcing translations of Section~\ref{sec:bridge}
formalizes only the constant embedding. A full mechanization of the
type-level translation, against an explicit forcing structure and
with the propositional rules verified to match our \texttt{iProp},
is the natural next step.

The abstract framework of Section~\ref{sec:abstract} parametrizes the
propositional layer over arbitrary well-founded forcing structures,
but the IMP and $\lambda$-calculus developments of
Sections~\ref{sec:imp}--\ref{sec:lam} are presented at the specific
$(\omega, \le)$ instance for clarity. Lifting them to be parametric
in the forcing structure is straightforward and not done here.

\paragraph{Future directions.}
Several extensions suggest themselves.

\begin{itemize}

\item \emph{Mechanizing the full JTS translation} (or a useful
fragment) and verifying that its propositional fragment coincides
definitionally with our \texttt{iProp} at the corresponding forcing
structure. This would convert the heuristic correspondence of
Section~\ref{sec:bridge} into a theorem.

\item \emph{Step-indexing over richer forcing structures} ---
transfinite ordinals (cf.\ Transfinite Iris
\cite{spies2021transfinite}), branching conditions for genuinely
concurrent step counters, or mixed structures combining several
step-index dimensions. The abstract framework already supports these
in principle; the work is to exhibit non-trivial verifications at
the new instances.

\item \emph{Lifting the Hoare-logic developments to the abstract
framework}, so that IMP and the $\lambda$-calculus admit
verifications parametric in the forcing structure. This is
straightforward but bookkeeping-heavy; it would make the parametric
story of Section~\ref{sec:abstract} the primary presentation rather
than an afterthought.

\item \emph{An integration with separation logic} along the lines
of Iris, lifting the forcing reading from the propositional logic
to the resource-aware extension. This is a substantial undertaking
but would let the forcing perspective speak directly to
state-of-the-art verification.

\end{itemize}

The broader point we hope the paper makes is that step-indexed
reasoning, viewed as a forcing construction, sits at the
intersection of three research traditions that do not usually talk
to each other: set-theoretic forcing, modal logic of provability,
and step-indexed semantics for verification. Each tradition contains
piece-of-the-picture insights that the other two could profit from.
The mechanization that accompanies this paper is intended both as
evidence for the technical claims and as a starting point for
others wishing to explore the intersection further.

%% =====================================================================
\bibliographystyle{ACM-Reference-Format}
\bibliography{refs}

\end{document}
