\section{Hoare Logic for IMP}
\label{sec:imp}

With the propositional layer in hand we can build a Hoare logic for
a simple imperative language.  We do this in the standard
calibrating setting --- a small while-language with mutable state,
which we will call IMP --- because the conceptual point of this
section is most cleanly visible there.  IMP has no higher-order
features, no allocation, and no concurrency, so the step indices we
attach to assertions are not yet doing technical work.  What the
section accomplishes is showing how the forcing-style propositional
layer of Section~\ref{sec:prop} lifts to a sound Hoare logic, and ---
crucially for what follows --- proving the \texttt{while} rule both
by meta-level induction on the step index and by an explicit appeal
to the internal Löb rule of Section~\ref{sec:prop}.  These two
proofs establish the same theorem, and exhibiting them side by side
makes precise the slogan that ``Löb's rule \emph{is} induction on the
step index''.

\paragraph{IMP and its semantics.}
The language is standard: arithmetic expressions over an integer
state, boolean expressions, and the usual control-flow commands.
Variables are natural numbers and the state is a function
\texttt{var -> Z}.

\begin{lstlisting}
Inductive cmd : Type :=
  | CSkip   : cmd
  | CAssign : var -> aexp -> cmd
  | CSeq    : cmd -> cmd -> cmd
  | CIf     : bexp -> cmd -> cmd -> cmd
  | CWhile  : bexp -> cmd -> cmd.
\end{lstlisting}

We give a small-step operational semantics
\texttt{step : cmd * state -> cmd * state -> Prop} with the usual
rules: assignment updates the state and reduces to \texttt{CSkip};
sequencing reduces its left component if not yet \texttt{CSkip} and
otherwise advances to its right component; conditionals branch on
the boolean evaluation; \texttt{while} either unfolds to a sequence
or terminates as \texttt{CSkip}.  Each step in this semantics
corresponds to one decrement of the step-index clock that the wp
predicate is about to track.

\paragraph{Step-indexed weakest precondition.}
The wp predicate for a command is a step-indexed analogue of the
classical Dijkstra wp: $\mathtt{wp\_at}\, n\, c\, Q\, s$ says that if
we run $c$ from state $s$ for at most $n$ small steps, no
intermediate configuration is bad and any final state satisfies
the postcondition $Q$.  Because the only ``bad'' configuration in
IMP is one that is neither \texttt{CSkip} nor able to step --- and
in IMP every non-\texttt{CSkip} command can step --- the safety
condition is automatic, and we do not need to record it
explicitly.  This invisibility of safety in IMP is what makes the
section a calibration; in Section~\ref{sec:lam} we will see what
happens when the language is rich enough for stuck terms to exist.

\begin{lstlisting}
Fixpoint wp_at (n : nat) (c : cmd)
               (Q : state -> Prop) (s : state) : Prop :=
  match n with
  | O => True
  | S k =>
      (c = CSkip -> Q s) /\
      (forall c' s', step (c, s) (c', s') -> wp_at k c' Q s')
  end.
\end{lstlisting}

The predicate is downward closed in $n$ (\texttt{wp\_at\_mono}) and
contravariantly monotone in the postcondition $Q$
(\texttt{wp\_at\_post\_mono}); both proofs are routine inductions on
$n$.  Lifting to an \texttt{iProp} is then immediate:

\begin{lstlisting}
Definition wp (c : cmd) (Q : state -> Prop) (s : state) : iProp :=
  MkProp (fun n => wp_at n c Q s) (wp_at_mono c Q s).
\end{lstlisting}

\paragraph{The Hoare triple.}
A Hoare triple is the \texttt{iProp} that says ``for every state
satisfying $P$, the wp of $c$ targeting $Q$ holds''.  By the
universal quantifier, this is again a downward-closed predicate in
$n$:

\begin{lstlisting}
Definition hoare (P : state -> Prop) (c : cmd)
                 (Q : state -> Prop) : iProp :=
  MkProp (fun n => forall s, P s -> wp_at n c Q s)
         (hoare_at_mono P c Q).

Definition hoare_valid P c Q : Prop := ⊤ ⊢ hoare P c Q.
\end{lstlisting}

We say the triple is \emph{valid} when it is forced at every
condition: $\top \vdash \mathtt{hoare}\, P\, c\, Q$, which by
unfolding is exactly the standard partial-correctness statement
``for every $n$ and every state $s$ with $P\, s$, the
$n$-step-bounded wp holds''.  This equivalence is the lemma
\texttt{hoare\_valid\_alt} in the development; we will use it freely
to move between the iProp-internal and the meta-level views of a
Hoare triple.

\paragraph{Soundness of the structural rules.}
The standard structural rules of partial-correctness Hoare logic
become small lemmas about \texttt{wp\_at}.  We state them in the
validity form:

\begin{lstlisting}
Lemma hoare_skip P : hoare_valid P CSkip P.
Lemma hoare_assign Q x a :
  hoare_valid (asgn_pre Q x a) (CAssign x a) Q.
Lemma hoare_seq P c1 R c2 Q :
  hoare_valid P c1 R ->
  hoare_valid R c2 Q ->
  hoare_valid P (CSeq c1 c2) Q.
Lemma hoare_if P b c1 c2 Q :
  hoare_valid (fun s => P s /\ beval s b = true) c1 Q ->
  hoare_valid (fun s => P s /\ beval s b = false) c2 Q ->
  hoare_valid P (CIf b c1 c2) Q.
Lemma hoare_consequence P P' c Q Q' :
  (forall s, P s -> P' s) -> hoare_valid P' c Q' ->
  (forall s, Q' s -> Q s) -> hoare_valid P c Q.
\end{lstlisting}

Each proof is a routine unfolding of the wp predicate at $n+1$,
followed by inversion on the available step.  Sequencing requires a
small auxiliary lemma \texttt{wp\_seq} that decomposes the wp of
\texttt{CSeq c1 c2} as the wp of \texttt{c1} targeting the wp of
\texttt{c2}; this is the only point where the step index moves
between the outer and inner wp, and a use of \texttt{wp\_at\_mono} is
needed to align the indices.  None of this requires the propositional
layer of Section~\ref{sec:prop}; we could have written the same five
rules entirely at the meta level, with $\mathtt{hoare\_valid}$ as a
plain ``forall $n, s$'' statement.  The propositional layer is
present because it will pay off in the next part of this section ---
and again, more decisively, in Section~\ref{sec:lam}.

\paragraph{The \texttt{while} rule: meta-level induction.}
The interesting rule is \texttt{while}.  Soundness of
\begin{center}
\(\dfrac{\{P \wedge b\}\, c\, \{P\}}{\{P\}\, \mathtt{while}\, b\, \mathtt{do}\, c\, \{P \wedge \neg b\}}\)
\end{center}
is, in step-indexed terms, the statement that a sound loop body
gives a sound loop --- but the loop's wp at $n+1$ unfolds into a
proof obligation about the loop's wp at $n$, which is exactly the
recursive structure we have been preparing for.  The straightforward
proof is by induction on $n$:

\begin{lstlisting}
Lemma hoare_while P b c :
  hoare_valid (fun s => P s /\ beval s b = true) c P ->
  hoare_valid P (CWhile b c)
              (fun s => P s /\ beval s b = false).
Proof.
  intros HC.
  rewrite hoare_valid_alt in HC.
  apply hoare_valid_alt.
  intros n. induction n as [|n IH]; intros s HP.
  - simpl. trivial.
  - rewrite wp_at_S. split.
    + discriminate.
    + intros c' s' Hstep. inversion Hstep; subst.
      * (* StepWhileTrue *)
        apply wp_seq.
        eapply wp_at_post_mono.
        2: { apply HC; split; [exact HP | assumption]. }
        intros s'' HP'. apply IH. exact HP'.
      * (* StepWhileFalse *)
        ... (* CSkip with the postcondition *)
Qed.
\end{lstlisting}

The structure is what one would write in any step-indexed Hoare
logic.  The case analysis on $n$ contributes a trivial base case
(the wp at $0$ is \texttt{True}) and an inductive step.  In the
\texttt{StepWhileTrue} case, the loop reduces to
$c ; \mathtt{while}\, b\, c$ at step $n$, and the proof obligation
for the remaining $\mathtt{while}\, b\, c$ piece is discharged by
the induction hypothesis $\mathtt{IH}$.

\paragraph{The same rule, via the internal Löb.}
There is also a proof of \texttt{hoare\_while} that does not use any
explicit induction on $n$ --- it appeals to the internal Löb rule of
Section~\ref{sec:prop} directly.  To use \texttt{iLob}, we need to
express the goal as an entailment $(\triangleright \varphi \to
\varphi) \vdash \varphi$.  This is done by an application of
\texttt{ientails\_trans}: we factor the desired entailment
$\top \vdash \mathtt{hoare}\, P\, (\mathtt{CWhile}\, b\, c)\, Q$
through the intermediate iProp $\triangleright \varphi \to \varphi$
for $\varphi := \mathtt{hoare}\, P\, (\mathtt{CWhile}\, b\, c)\, Q$,
prove the intermediate $\top \vdash \triangleright \varphi \to
\varphi$ directly, and finish with \texttt{apply iLob}:

\begin{lstlisting}
Lemma hoare_while_iLob P b c :
  hoare_valid (fun s => P s /\ beval s b = true) c P ->
  hoare_valid P (CWhile b c)
              (fun s => P s /\ beval s b = false).
Proof.
  intros HC. rewrite hoare_valid_alt in HC.
  unfold hoare_valid.
  apply ientails_trans with
    (iLater (hoare P (CWhile b c) (...)) →
     hoare P (CWhile b c) (...)).
  - (* ⊤ ⊢ ▷ φ → φ *)
    intros n _ m Hmn Hlater s HP.
    destruct m as [|m'].
    + simpl. trivial.
    + simpl in Hlater. rewrite wp_at_S. split.
      * discriminate.
      * intros c' s' Hstep. inversion Hstep; subst.
        -- (* StepWhileTrue *)
           apply wp_seq.
           eapply wp_at_post_mono.
           2: { apply HC; split; [exact HP | assumption]. }
           intros s'' HP''. apply Hlater. exact HP''.
        -- (* StepWhileFalse *)
           ...
  - apply iLob.
Qed.
\end{lstlisting}

The body of the first bullet --- the proof that $\top$ entails
$\triangleright \varphi \to \varphi$ at every condition --- contains
no induction on $n$.  We case-split on the depth $m$ obtained from
the iImpl, find $\triangleright \varphi$ supplied to us at depth
$m$, and in the $\mathtt{S}\, m'$ case use it (it is $\varphi$ at
depth $m'$) to discharge the recursive obligation.  All of the
recursive content is hidden inside \texttt{iLob}'s own proof, which
of course does eventually go by induction on $n$ --- but
\emph{once}, in Section~\ref{sec:prop}, and not at every Hoare-style
soundness proof.

\paragraph{The packaged soundness theorem.}
With both proofs available, we package the six structural rules into
a single theorem statement:

\begin{lstlisting}
Theorem imp_hoare_rules :
  (forall P, hoare_valid P CSkip P) /\
  (forall Q x a, hoare_valid (asgn_pre Q x a) (CAssign x a) Q) /\
  (forall P c1 R c2 Q, ... -> ... -> ...) /\  (* seq *)
  (forall P b c1 c2 Q, ... -> ... -> ...) /\  (* if *)
  (forall P b c, ... -> ...) /\              (* while *)
  (forall P P' c Q Q', ... -> ... -> ... -> ...).  (* consequence *)
\end{lstlisting}

This is the partial-correctness soundness statement for Hoare logic
over IMP, in the forcing-style semantics.  It is structurally
identical to the soundness statement one would prove for IMP using
the standard direct semantics, and that identity is part of the
point: at the first-order level the forcing reading is a different
\emph{presentation} of the same theorems.  In
Section~\ref{sec:lam} we will move to a setting where the
re-presentation produces a noticeably different proof obligation.

