\section{An Abstract Forcing Framework}
\label{sec:abstract}

So far, the forcing notion has been the specific poset
$(\omega, \le)$.  The proofs of Sections~\ref{sec:prop}--\ref{sec:lam}
have used the natural numbers throughout: as the type of conditions,
as the index of the wp predicate, and as the well-founded order on
which Löb's rule was proved.  But none of these uses required $\omega$
specifically.  In this section we abstract the framework over an
arbitrary well-founded forcing notion, and prove Löb's rule \emph{once}
at this level of generality.

The abstraction makes precise two claims that have been hovering
over the development.  First, the choice of $\omega$ in
Section~\ref{sec:prop} is a presentation convenience rather than a
foundational commitment: any well-founded preorder will do.  Second,
the relationship between Löb's rule and well-founded induction that
we have been describing analogically becomes a literal identity in
the abstract setting --- the abstract \texttt{iLob} is proved by
\texttt{apply (well\_founded\_ind fs\_lt\_wf)}.

\paragraph{The forcing structure.}
We package the assumed structure into a record:

\begin{lstlisting}
Record ForcingStructure : Type := MkFS {
  fs_cond : Type;
  fs_le : fs_cond -> fs_cond -> Prop;
  fs_lt : fs_cond -> fs_cond -> Prop;
  fs_le_refl : forall p, fs_le p p;
  fs_le_trans : forall p q r,
                  fs_le p q -> fs_le q r -> fs_le p r;
  fs_lt_le : forall p q, fs_lt p q -> fs_le p q;
  fs_lt_le_lt : forall p q r,
                  fs_lt p q -> fs_le q r -> fs_lt p r;
  fs_lt_wf : well_founded fs_lt
}.
\end{lstlisting}

A forcing structure is a preorder $(\mathtt{fs\_cond}, \mathtt{fs\_le})$ together with a strict relation $\mathtt{fs\_lt}$ contained in $\mathtt{fs\_le}$ and compatible with $\mathtt{fs\_le}$ on the right, such that $\mathtt{fs\_lt}$ is well-founded.  The compatibility axiom $\mathtt{fs\_lt\_le\_lt}$ is the technical ingredient that makes the later modality well-defined in the abstract setting; we explain this below.

\paragraph{iProp, connectives, and \texttt{iLater} at the abstract level.}
Over an arbitrary forcing structure $\mathtt{FS}$, the \texttt{iProp} record is the obvious adaptation: a downward-closed predicate on $\mathtt{fs\_cond}\, \mathtt{FS}$ with respect to $\mathtt{fs\_le}\, \mathtt{FS}$.  The propositional connectives are defined exactly as in Section~\ref{sec:prop}, with the order $\le$ replaced by $\mathtt{fs\_le}\, \mathtt{FS}$.  The interesting case is the later modality.  At the specific $(\omega, \le)$ instance, we defined $\triangleright \varphi$ at depth $n$ by a match on the structure of $n$ (\texttt{True} if $n = 0$, $\varphi$ at $n - 1$ otherwise).  At an arbitrary forcing notion, we cannot match on conditions because the carrier is not a constructive inductive type.  We instead give the modality a uniform definition: $\triangleright \varphi$ holds at a condition $p$ if and only if $\varphi$ holds at every strict predecessor of $p$:

\begin{lstlisting}
Definition iLater (φ : iProp) : iProp :=
  MkProp
    (fun p => forall p', fs_lt FS p' p -> p' ⊩ φ)
    (iLater_mono_helper φ).
\end{lstlisting}

\noindent
At a \emph{minimal} condition --- one with no strict predecessors ---
this is vacuously \texttt{True}.  At a successor condition, it says
the proposition holds at every preceding stage.  The downward-closure
monotonicity of $\triangleright \varphi$ requires the
$\mathtt{fs\_lt\_le\_lt}$ axiom to chain a strict predecessor of $q$
through a $\le$-arrow $q \le p$ into a strict predecessor of $p$.

The new definition agrees with the Section~\ref{sec:prop} version
when we instantiate at $(\omega, \le)$.  At depth $0$, both definitions give \texttt{True}.  At depth $n+1$, the new definition says ``$\varphi$ holds at every $m < n+1$'', which by downward closure of $\varphi$ is equivalent to ``$\varphi$ holds at $n$'', exactly the old definition.

\paragraph{Löb's rule, via well-founded induction.}
The internal Löb rule, in the abstract framework, is a direct
appeal to \texttt{well\_founded\_ind} on $\mathtt{fs\_lt}$:

\begin{lstlisting}
Lemma iLob (φ : iProp) : (▷ φ → φ) ⊢ φ.
Proof.
  intros p. revert p.
  apply (well_founded_ind (fs_lt_wf FS)
                          (fun p => p ⊩ (▷ φ → φ) -> p ⊩ φ)).
  intros p IH Hp.
  apply Hp; [apply (fs_le_refl FS)|].
  intros p' Hp'.
  apply IH; [exact Hp'|].
  eapply imono; [apply (fs_lt_le FS); exact Hp' | exact Hp].
Qed.
\end{lstlisting}

The proof is almost mechanical given the definitions.  We do
well-founded induction on $p$; at each $p$, the induction hypothesis
says that for every strict predecessor $p'$ of $p$, the implication
$\triangleright \varphi \to \varphi$ at $p'$ already gives us $\varphi$
at $p'$.  We then use $\mathtt{Hp}$ at $p$ itself --- via the
$\mathtt{iImpl}$'s Kripke condition with $\mathtt{fs\_le\_refl}$ to
witness $p \le p$ --- which leaves us to prove $\triangleright \varphi$
at $p$.  By the definition of $\triangleright$, this is the
quantified statement that $\varphi$ holds at every strict predecessor
$p'$; for each such $p'$, the induction hypothesis applies (we have
$\mathtt{Hp}$ at $p'$ by downward closure).

This is the precise statement of the equivalence we have been
gesturing at: \emph{Löb's rule is well-founded induction on the
forcing order, internalized into the logic}.  The metatheoretic
content is the same; the proof at the meta level uses
\texttt{well\_founded\_ind}, and the proof at the iProp level uses
\texttt{iLob}.

\paragraph{Instances.}
We exhibit two concrete forcing structures.  The first is the
natural numbers with their usual order:

\begin{lstlisting}
Definition nat_FS : ForcingStructure :=
  {| fs_cond := nat;
     fs_le := Nat.le;
     fs_lt := Nat.lt;
     fs_le_refl := Nat.le_refl;
     fs_le_trans := Nat.le_trans;
     fs_lt_le := Nat.lt_le_incl;
     fs_lt_le_lt := Nat.lt_le_trans;
     fs_lt_wf := lt_wf
  |}.
\end{lstlisting}

\noindent
At $\mathtt{nat\_FS}$, the abstract framework recovers
Section~\ref{sec:prop}.  The check $\mathtt{Check}\, \mathtt{iLob}\, \mathtt{nat\_FS}$ confirms the instantiation: the abstract
$\mathtt{iLob}$ applied to $\mathtt{nat\_FS}$ has the type we expect,
namely an entailment about Phase 1's $\mathtt{iLater}$ and
$\mathtt{iImpl}$ (specialized to the natural-number instance) holding
in our Phase 1's $\mathtt{iProp}$.

The second instance is a non-trivially-different forcing notion: the
pointwise product $\omega \times \omega$.

\begin{lstlisting}
Definition le_pair (p q : nat * nat) : Prop :=
  fst p <= fst q /\ snd p <= snd q.

Definition lt_pair (p q : nat * nat) : Prop :=
  le_pair p q /\ ~ (fst p = fst q /\ snd p = snd q).
\end{lstlisting}

\noindent
The order $\le_{\mathrm{pw}}$ is the componentwise $\le$; the strict
order $<_{\mathrm{pw}}$ is the obvious one (componentwise $\le$ with
inequality in at least one component).  This poset is not totally
ordered: the pairs $(1, 0)$ and $(0, 1)$ are incomparable.  We
verify well-foundedness of $<_{\mathrm{pw}}$ by exhibiting the measure
$\mathtt{fst}\, p + \mathtt{snd}\, p$ into $\mathbb{N}$, and using
the standard library lemma $\mathtt{well\_founded\_lt\_compat}$:

\begin{lstlisting}
Lemma lt_pair_wf : well_founded lt_pair.
Proof.
  apply (well_founded_lt_compat _
           (fun p => fst p + snd p)).
  intros p q [[Hf Hs] Hneq].
  destruct p, q; simpl in *.
  destruct (Nat.eq_dec n n1).
  - subst. destruct (Nat.eq_dec n0 n2).
    + subst. exfalso. apply Hneq. split; reflexivity.
    + lia.
  - lia.
Qed.

Definition prod_FS : ForcingStructure :=
  {| fs_cond := nat * nat; ... |}.
\end{lstlisting}

\noindent
At $\mathtt{prod\_FS}$, the framework provides a step-indexed logic
with \emph{two independent} step counters.  A natural application is
to model two parallel components that step independently --- the
step index measures observation depth in each component
separately, and Löb's rule lets one reason about recursion that
descends in either or both components.  We do not develop a Hoare
logic over $\mathtt{prod\_FS}$ in this paper, but the existence of
the instance is conceptually important: it shows that the framework
is not tied to ordinal step-indices in particular, and is in
particular applicable to the kind of branching well-founded
structure one expects in concurrent settings.

\paragraph{What is recovered.}
At the abstract level, we have shown that:

\begin{enumerate}

\item the data of a step-indexed logic is captured by a
\texttt{ForcingStructure} record;

\item the propositional fragment of step-indexed Hoare logic, with
the Heyting connectives and the later modality, is parametric in
this data;

\item the internal Löb rule, in any such structure, is a direct
instance of well-founded induction on the strict refinement
relation; and

\item the standard $(\omega, \le)$ choice, used throughout
Sections~\ref{sec:prop}--\ref{sec:lam} and throughout the modern
step-indexed literature, is one instance, with at least one
non-trivially-different alternative ($\omega \times \omega$ with the
pointwise order).

\end{enumerate}

\noindent
The IMP and $\lambda$-calculus developments of
Sections~\ref{sec:imp}--\ref{sec:lam} could be lifted to be
parametric in the forcing structure as well, with mostly
bookkeeping changes; we have not done so here, for the sake of
readability.

