\section{Conclusion}
\label{sec:conclusion}

We have given a forcing-style semantics for step-indexed Hoare logic,
mechanized in Rocq across the propositional layer, an IMP Hoare
logic, and an untyped $\lambda$-calculus with general recursion. The
forcing reading places the step-indexed technique in direct
correspondence with Cohen's ramified forcing construction: the
natural numbers are the forcing conditions, the later modality
$\triangleright$ is the level-descent operator, and Löb's rule is
well-founded induction on the forcing order, internalized into the
logic.

\paragraph{What we found.}
Two observations stand out. First, in the first-order setting of
IMP, the \texttt{while} rule admits two structurally identical
soundness proofs: an outer induction on the step index, and an
appeal to the internal Löb rule. The two proofs have the same
content because at the meta level, Löb's rule \emph{is} induction on
the step index. Mechanizing both side by side, as we do, makes the
equivalence concrete rather than analogical.

Second, in the higher-order setting of the untyped $\lambda$-calculus
with \texttt{fix}, the analogous soundness proof has a different
shape. The soundness predicate for a recursive function is itself
defined in terms of the function being verified, and unrolling this
self-reference at the meta level produces a non-trivial induction
that does not arise from the program's step count alone. The natural
proof, which we mechanize, invokes the internal Löb rule directly:
``show that under the assumption $\triangleright \varphi$ we have
$\varphi$, then conclude $\varphi$ by Löb.'' No meta-level induction
on a step count is needed. The structural difference between the IMP
and $\lambda$-fix proofs is a difference in the
\emph{recursion locus} of the soundness predicate
(Section~\ref{sec:imp-vs-lam}): for IMP, the predicate is recursive
only in the step index, and outer induction on $n$ matches its
structure; for $\lambda$-fix, the predicate is recursive in the
program itself, and internal Löb is the natural tool that plays to
that structure.

\paragraph{Three further observations.}
The abstract framework of Section~\ref{sec:abstract} makes precise
that the choice of $\omega$ as the step-index domain is a
\emph{computational} convenience rather than a foundational one. Any
well-founded forcing notion supports the same internal logic, with
$\triangleright$ defined uniformly as ``forces at every strict
predecessor'' and Löb derived by well-founded induction. We exhibit
the pointwise product $\omega \times \omega$ as a second instance, but
the framework is generic.

Section~\ref{sec:lob-essential} shows that the well-foundedness
cannot be dropped: we construct a non-well-founded forcing notion
over $\mathbb{Z}$ in which Löb's rule fails. This is a special case
of the model theory of \textsf{GL}, but the statement for step-indexed
Hoare logic does not, as far as we are aware, appear in the
verification literature. The claim has methodological weight:
step-indexed Hoare logic is not just ``Hoare logic with a step
counter''; it requires a well-founded forcing notion to be sound at
all.

Section~\ref{sec:bridge} verifies that the iProp framework is the
propositional fragment of the Jaber--Tabareau--Sozeau forcing
translation of CIC, via a constant embedding
$\mathsf{Prop} \hookrightarrow \mathtt{iProp}$ that we prove to be a
Heyting algebra homomorphism. The $\triangleright$ modality is the
genuinely new content of iProp over the embedded image, and arises
uniformly from the well-founded strict refinement on conditions.

\paragraph{Limitations.}
The Hoare logic for the $\lambda$-calculus in
Section~\ref{sec:lam} is, at present, restricted to partial
correctness against a single specification predicate $S$ that
serves as both pre- and post-condition. The fix soundness rule
\texttt{wp\_fix} (Section~\ref{sec:lam}) accordingly handles only
this case. A more general rule with distinct pre/post-conditions, a
rule for fix at higher types, and an integration with separation
logic on a heap are all natural directions but lie outside the
present development.

The bridge to JTS forcing translations of Section~\ref{sec:bridge}
formalizes only the constant embedding. A full mechanization of the
type-level translation, against an explicit forcing structure and
with the propositional rules verified to match our \texttt{iProp},
is the natural next step.

The abstract framework of Section~\ref{sec:abstract} parametrizes the
propositional layer over arbitrary well-founded forcing structures,
but the IMP and $\lambda$-calculus developments of
Sections~\ref{sec:imp}--\ref{sec:lam} are presented at the specific
$(\omega, \le)$ instance for clarity. Lifting them to be parametric
in the forcing structure is straightforward and not done here.

\paragraph{Future directions.}
Several extensions suggest themselves.

\begin{itemize}

\item \emph{Mechanizing the full JTS translation} (or a useful
fragment) and verifying that its propositional fragment coincides
definitionally with our \texttt{iProp} at the corresponding forcing
structure. This would convert the heuristic correspondence of
Section~\ref{sec:bridge} into a theorem.

\item \emph{Step-indexing over richer forcing structures} ---
transfinite ordinals (cf.\ Transfinite Iris
\cite{spies2021transfinite}), branching conditions for genuinely
concurrent step counters, or mixed structures combining several
step-index dimensions. The abstract framework already supports these
in principle; the work is to exhibit non-trivial verifications at
the new instances.

\item \emph{Lifting the Hoare-logic developments to the abstract
framework}, so that IMP and the $\lambda$-calculus admit
verifications parametric in the forcing structure. This is
straightforward but bookkeeping-heavy; it would make the parametric
story of Section~\ref{sec:abstract} the primary presentation rather
than an afterthought.

\item \emph{An integration with separation logic} along the lines
of Iris, lifting the forcing reading from the propositional logic
to the resource-aware extension. This is a substantial undertaking
but would let the forcing perspective speak directly to
state-of-the-art verification.

\end{itemize}

The broader point we hope the paper makes is that step-indexed
reasoning, viewed as a forcing construction, sits at the
intersection of three research traditions that do not usually talk
to each other: set-theoretic forcing, modal logic of provability,
and step-indexed semantics for verification. Each tradition contains
piece-of-the-picture insights that the other two could profit from.
The mechanization that accompanies this paper is intended both as
evidence for the technical claims and as a starting point for
others wishing to explore the intersection further.

