\section{Hoare Logic for an Untyped $\lambda$-Calculus with Recursion}
\label{sec:lam}

For IMP, step-indexing was overkill: the wp predicate could have been
defined without indexing at all, and the index was a presentation
convenience.  We now move to a language where step-indexing is no
longer optional --- an untyped $\lambda$-calculus with a fix-point
operator.  Two things change.  First, the wp predicate must
explicitly encode \emph{safety}: applying an integer as a function,
or branching on a lambda, leaves the computation stuck, and we want
the wp to rule such configurations out rather than treat them as
vacuously good.  Second, the recursion introduced by the fix-point
operator turns the soundness predicate for a recursive function into
something defined in terms of itself, and the natural soundness
proof for the corresponding Hoare rule no longer goes through
meta-level induction on the step index.  It goes through an
appeal to the internal Löb rule of Section~\ref{sec:prop}, used
directly, with no outer induction on $n$.

\paragraph{The language.}
We use a de-Bruijn representation, with integer constants for the
sake of having a non-trivial base type, an \texttt{if0} conditional
that branches on whether its argument equals zero, $\lambda$-abstraction
and application, and an explicit fix-point binder:

\begin{lstlisting}
Inductive expr : Type :=
  | EVar  : nat -> expr
  | EInt  : Z -> expr
  | EPlus : expr -> expr -> expr
  | EIf0  : expr -> expr -> expr -> expr
  | ELam  : expr -> expr
  | EApp  : expr -> expr -> expr
  | EFix  : expr -> expr.

Inductive value : expr -> Prop :=
  | VInt : forall n, value (EInt n)
  | VLam : forall body, value (ELam body)
  | VFix : forall body, value (EFix body).
\end{lstlisting}

The fix-point form $\mathtt{EFix}\, body$ binds the self-reference
at de-Bruijn index $0$ inside $body$; we treat it as a value so that
it can be passed as an argument, and we let the self-unfolding fire
only when the fix is \emph{applied}.  Substitution is the standard
single substitution; we omit its definition here.  The small-step
relation includes the usual rules for arithmetic and conditionals
together with $\beta$-reduction
\(\mathtt{EApp}\, (\mathtt{ELam}\, body)\, v \to body[v / 0]\)
and the fix-unfolding rule

\begin{lstlisting}
| StepFix : forall body v, value v ->
    EApp (EFix body) v ~>
    EApp (subst (EFix body) body) v.
\end{lstlisting}

\noindent
which substitutes the self-reference $\mathtt{EFix}\, body$ for index
$0$ inside the body before letting an ordinary $\beta$-reduction
proceed.  All reductions are call-by-value.

\paragraph{Step-indexed wp, with safety.}
The wp predicate is similar in shape to the IMP version but has an
additional conjunct that pins down safety.  At index $n+1$ we ask
that the expression either is a value satisfying the postcondition,
or is reducible and every reduct still satisfies the wp at index
$n$:

\begin{lstlisting}
Fixpoint wp_at (n : nat) (e : expr)
               (Q : expr -> Prop) : Prop :=
  match n with
  | O => True
  | S k =>
      (value e -> Q e) /\
      (value e \/ exists e', step e e') /\
      (forall e', step e e' -> wp_at k e' Q)
  end.
\end{lstlisting}

The middle conjunct (\texttt{value e \textbackslash{}/ exists e', step e e'}) is the new ingredient.  Without it, a stuck non-value expression such as \texttt{EApp (EInt 5) (EInt 3)} would satisfy the wp vacuously: the first conjunct is discharged (the expression is not a value), and the third quantifies over a step relation that has no inhabitants for the stuck expression.  With it, the stuck expression fails the wp.  This is the price of working with a language whose syntax admits ill-typed combinations; the IMP development of Section~\ref{sec:imp} did not need it because every non-\texttt{CSkip} command can step.

\paragraph{Basic rules.}
Two general structural rules carry every soundness proof.  The first
says that a value satisfies the wp targeting any postcondition it
semantically satisfies:

\begin{lstlisting}
Lemma wp_value v Q :
  value v -> Q v -> wp_valid v Q.
\end{lstlisting}

\noindent
The second is the \emph{pure step} rule: if every reduct of $e$ is
$e'$ and $e'$ satisfies the wp, then so does $e$.

\begin{lstlisting}
Lemma wp_pure_step_at n e e' Q :
  ~ value e ->
  (exists e'', step e e'') ->
  (forall e'', step e e'' -> e'' = e') ->
  wp_at n e' Q ->
  wp_at (S n) e Q.
\end{lstlisting}

\noindent
The pure-step rule is what propagates the wp through any
$\beta$-reduction or fix-unfolding step.  In particular, it is the
device by which we will discharge the fix-unfolding step in the
\texttt{wp\_fix} soundness proof below.

We do not develop dedicated wp rules for \texttt{EPlus}, \texttt{EIf0}, or \texttt{EApp} at this stage --- they either need an evaluation-context framework or unfold case-by-case via \texttt{wp\_pure\_step}.  Neither addition is necessary for the conceptual point of this section; we will see them play out concretely in the worked example below.

\paragraph{The specification predicate for recursive functions.}
To state a meaningful soundness rule for \texttt{fix}, we need a
shape of specification appropriate to a recursive function.  The
simplest non-trivial shape is the \emph{closed} spec, where the same
predicate $S$ on values plays both the role of a precondition on the
argument and the role of a postcondition on the result.  We package
this as an iProp:

\begin{lstlisting}
Definition recursive_spec (e : expr) (S : expr -> Prop) : iProp :=
  MkProp
    (fun n => forall v, value v -> S v ->
                          wp_at n (EApp e v) S)
    (recursive_spec_mono e S).
\end{lstlisting}

\noindent
The proposition $\mathtt{recursive\_spec}\, e\, S$ at depth $n$ says:
``for every value $v$ satisfying $S$, applying $e$ to $v$ is safe
through $n$ steps, and any value it produces satisfies $S$''.
Crucially, $S$ appears \emph{twice} --- once as the precondition,
once as the postcondition.  This is the source of the
self-reference: to prove the spec for $\mathtt{EFix}\, body$ at
depth $n+1$, we need to know the spec for $\mathtt{EFix}\, body$ at
depth $n$ in order to handle the recursive call inside the body.

\paragraph{The \texttt{wp\_fix} soundness rule.}
The soundness rule we prove is:
\begin{theorem}[\texttt{wp\_fix}, the soundness rule for fix-points]
Let \texttt{body} be an expression and $S$ a predicate.  Suppose
that for every value $e_{\mathrm{self}}$ for which
$\mathtt{recursive\_spec}\, e_{\mathrm{self}}\, S$ holds, we also
have $\mathtt{recursive\_spec}\, (\mathtt{subst}\, e_{\mathrm{self}}\, body)\, S$ (a Lipschitz-like assumption on the body).
Then $\top \vdash \mathtt{recursive\_spec}\, (\mathtt{EFix}\, body)\, S$.
\end{theorem}

In words: if the body is a well-behaved transformer of recursive
specs --- if it produces a function satisfying the spec from a
self-reference satisfying the spec --- then the fix-point itself
satisfies the spec.

The natural proof uses the internal Löb rule directly.  We factor
the goal $\top \vdash \mathtt{recursive\_spec}\, (\mathtt{EFix}\, body)\, S$ through the iProp $(\triangleright \varphi \to \varphi)$ for $\varphi := \mathtt{recursive\_spec}\, (\mathtt{EFix}\, body)\, S$, prove the intermediate
$\top \vdash \triangleright \varphi \to \varphi$ directly --- this
is a finite piece of step-indexed reasoning, with no induction on $n$ ---
and close with \texttt{apply iLob}:

\begin{lstlisting}
Theorem wp_fix body S :
  (forall e_self, value e_self ->
     recursive_spec e_self S ⊢ recursive_spec (subst e_self body) S) ->
  ⊤ ⊢ recursive_spec (EFix body) S.
Proof.
  intros Hbody.
  apply ientails_trans with
    (iLater (recursive_spec (EFix body) S) →
     recursive_spec (EFix body) S).
  - (* ⊤ ⊢ ▷ φ → φ *)
    intros n _ m Hmn Hlater.
    destruct m as [|m'].
    + intros v Hvalue HS. simpl. trivial.
    + simpl in Hlater. intros v Hvalue HS.
      apply (wp_pure_step_at m'
        (EApp (EFix body) v)
        (EApp (subst (EFix body) body) v) S).
      * apply not_value_app_fix.
      * apply step_fix_progress; exact Hvalue.
      * intros e' Hs. apply step_fix_det; auto.
      * apply (Hbody (EFix body) (VFix body) m'
                     Hlater v Hvalue HS).
  - apply iLob.
Qed.
\end{lstlisting}

It is worth spelling out what the proof actually does.  The
\texttt{apply ientails\_trans} step factors the goal through Löb's
form.  In the body of the first bullet, the case analysis on $m$
contributes only a trivial base case ($m = 0$, where the wp is
\texttt{True}) and an inductive step.  In the inductive case,
the hypothesis $\mathtt{Hlater}$ supplies us with
$\mathtt{recursive\_spec}\, (\mathtt{EFix}\, body)\, S$ at the
\emph{predecessor} index $m'$ --- which is exactly the data we
need to satisfy the precondition of $\mathtt{Hbody}$ when
discharging the recursive call.  The fix-unfolding step from
$\mathtt{EApp}\, (\mathtt{EFix}\, body)\, v$ is then a single
application of \texttt{wp\_pure\_step\_at}, which leaves us at
index $m'$ inside the substituted body --- and the substituted body
satisfies the spec at $m'$ by Hbody and Hlater.

There is no induction on $n$ in this proof.  The induction is
hidden inside \texttt{iLob}'s own proof, which we wrote once in
Section~\ref{sec:prop}.  This is the structural feature that we
want to spotlight in Section~\ref{sec:imp-vs-lam}: the soundness
predicate $\mathtt{recursive\_spec}$ is recursive in the function
being verified, and so the natural soundness proof must invoke
\emph{some} recursive reasoning principle.  In a meta-level
induction we would have to manufacture that recursion from outside;
the internal Löb rule has it built in.

\paragraph{Distinct pre- and postconditions.}
The single-predicate \texttt{recursive\_spec} is the cleanest shape
for stating the diagnostic of Section~\ref{sec:imp-vs-lam}, but it is
not the most general one.  In a standard Hoare-triple setting one
wants a precondition $P$ on the argument and a postcondition $Q$ on
the result, with the two distinct.  The mechanization includes the
generalised form:
\begin{lstlisting}
Definition recursive_spec_pp (e : expr) (P Q : expr -> Prop) : iProp :=
  MkProp (fun n => forall v, value v -> P v -> wp_at n (EApp e v) Q)
         (recursive_spec_pp_mono e P Q).

Theorem wp_fix_pp body P Q :
  (forall e_self, value e_self ->
     recursive_spec_pp e_self P Q ⊢
       recursive_spec_pp (subst e_self body) P Q) ->
  ⊤ ⊢ recursive_spec_pp (EFix body) P Q.
\end{lstlisting}
The proof of \texttt{wp\_fix\_pp} is structurally identical to the
proof of \texttt{wp\_fix}: the same factoring through
$\triangleright \varphi \to \varphi$ for the corresponding $\varphi$,
the same \texttt{wp\_pure\_step\_at} for the fix-unfolding step, the
same closing \texttt{apply iLob}.  The locus argument of
Section~\ref{sec:imp-vs-lam} carries over without change.  The
single-predicate form is what we use in the worked examples below,
because it makes the verifications shortest; the generalised form is
the one to reach for in actual program verification.

\paragraph{Worked examples.}
We close the section with two short verifications that exercise
the framework end-to-end.  The first is a fix-based identity ---
$\mathtt{recursive\_id} := \mathtt{EFix}\, (\mathtt{ELam}\, (\mathtt{EVar}\, 0))$, which observably equals the plain identity $\lambda x.x$ but forces a fix-unfolding step at every application:

\begin{lstlisting}
Definition recursive_id : expr := EFix (ELam (EVar 0)).

Theorem recursive_id_spec (S : expr -> Prop) :
  ⊤ ⊢ recursive_spec recursive_id S.
\end{lstlisting}

\noindent
The body $\mathtt{ELam}\, (\mathtt{EVar}\, 0)$ does not mention the
self-reference at index $1$, so substitution into it is the identity
on expressions, and the Lipschitz premise of \texttt{wp\_fix} becomes
a non-recursive statement that we discharge directly with
\texttt{wp\_pure\_step\_at}.  This example is the simplest possible
exercise of the rule: it confirms that the framework can verify a
recursive function whose recursive structure is trivial.

The second example is the canonical divergent fix-point:

\begin{lstlisting}
Definition bottom : expr := EFix (EVar 0).

Theorem bottom_spec (S : expr -> Prop) :
  ⊤ ⊢ recursive_spec bottom S.
Proof.
  apply wp_fix. intros e_self Hself.
  apply ientails_refl.
Qed.
\end{lstlisting}

\noindent
For $\mathtt{bottom}$, the body \emph{is} the bare self-reference at
index $0$, so substituting $e_{\mathrm{self}}$ into it gives
$e_{\mathrm{self}}$ unchanged, and the Lipschitz premise of
$\mathtt{wp\_fix}$ collapses to reflexivity.  The conclusion ---
that $\mathtt{bottom}$ satisfies every recursive spec --- is the
familiar observation that a divergent program satisfies any
partial-correctness specification vacuously; here it appears as a
one-liner because the forcing framework absorbs the divergence into
the wp's index hierarchy.

